### 2018-10-24 project_proposal/pp.tex
added project_proposal

\documentclass[11pt,a4paper]{scrartcl}
%{{{ general stuff
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[hidelinks]{hyperref}
\usepackage{float} % use H in figure placement

%}}}
%{{{ graphics
\usepackage{graphicx} % Bilder
%}}}
%{{{ math
\usepackage{mathrsfs} % mathcal and mathscr
\usepackage{mathtools, amssymb, amsthm}
\usepackage{bm} % cool bold symbols
%}}}

% a todo command===============================================================
\newcounter{todocounter}
\newcommand{\todo}[2][noisnotdefined]{
 \marginpar{\fcolorbox{black}{yellow}{\footnotesize\textbf{todo}}
 \ifthenelse{\equal{#1}{noisnotdefined}}{}{\textcolor{black}{\newline\tiny #1}}}
 \textbf{\ifthenelse{\equal{#2}{.}}
   {\fcolorbox{blue}{white}{\textcolor{blue}{$\maltese$}}}{{\textcolor{blue}{#2}}}}
 \refstepcounter{todocounter}}

%===============================================================================

\date{}

\begin{document}

\section*{Project Proposal for Genetic Algorithms and Evolutionary Programming
by Felix Bartel and D\` avid Kerekes}

Our idea was to apply a genetic algorithm to a neuronal network which is known as neuroevolution.
A big advantage of this sheme is that one only needs an indicator how good the neuronal network is given a certain task.
This can for instance be given by the outcome of a game.
Our game of choice was Blobby Volly 2 which is a simple two-dimensional vollyball game with open source written in C++ with lua support, cf. Figure \ref{fig:screenshot}.

\begin{figure}[H]
\center
\includegraphics[width=0.7\textwidth]{img/screenshot.png}
\caption{screenshot of the game}
\label{fig:screenshot}
\end{figure}

A scene from the game could be described by the $x$ and $y$ coordinates of the players and the ball, their corresponding velocities and the current number of ball contacts.
These arguments or a subset of them could then be used as input for the neuronal network.
The output would then be the players movement which only consists of going right, left or jumping.
Assuming the layers of the neuronal network have $n_1,\dots,n_L$ nodes we then have to optimize a function of $\sum_{l=1}^{L-1} n_ln_{l+1}$ variables if we fix the size of the network.
A few possible extensions are
\begin{itemize}
\item different crossover operators
\item differnet mutation operators (maybe use a covanriance matrix, if the size allows it)
\item make the size of the network variable
\item vary the input or output nodes
\item vary the opponent to train (maybe also the neuronal network itself)
\end{itemize}

\end{document}

### 2018-11-05 project/blobby-1.0_fast/config.h
second commit

@@ -1,0 +1,22 @@
+/*=============================================================================
+Blobby Volley 2
+Copyright (C) 2006 Jonathan Sieber (jonathan_sieber@yahoo.de)
+
+This program is free software; you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation; either version 2 of the License, or
+(at your option) any later version.
+
+This program is distributed in the hope that it will be useful,
+but WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+GNU General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with this program; if not, write to the Free Software
+Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
+=============================================================================*/
+
+#pragma once
+
+#define BLOBBY_INSTALL_PREFIX "/usr/local"

### 2018-11-05 project/blobby-1.0_fast/src/GameLogic.cpp

@@ -1,0 +1,946 @@
+/*=============================================================================
+Blobby Volley 2
+Copyright (C) 2006 Jonathan Sieber (jonathan_sieber@yahoo.de)
+Copyright (C) 2006 Daniel Knobe (daniel-knobe@web.de)
+
+This program is free software; you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation; either version 2 of the License, or
+(at your option) any later version.
+
+This program is distributed in the hope that it will be useful,
+but WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+GNU General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with this program; if not, write to the Free Software
+Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
+=============================================================================*/
+
+/* header include */
+#include "GameLogic.h"
+
+/* includes */
+#include <cassert>
+#include <cmath>
+#include <iostream>
+
+extern "C"
+{
+#include "lua/lua.h"
+#include "lua/lauxlib.h"
+#include "lua/lualib.h"
+}
+
+#include "FileRead.h"
+#include "GameLogicState.h"
+#include "DuelMatch.h"
+#include "GameConstants.h"
+#include "IUserConfigReader.h"
+#include "IScriptableComponent.h"
+#include "PlayerInput.h"
+
+
+int lua_toint(lua_State* state, int index)
+{
+	double value = lua_tonumber(state, index);
+	return int(value + (value > 0 ? 0.5 : -0.5));
+}
+
+
+/* implementation */
+
+/// how many steps must pass until the next hit can happen
+const int SQUISH_TOLERANCE = 11;
+
+const std::string FALLBACK_RULES_NAME = "__FALLBACK__";
+const std::string DUMMY_RULES_NAME = "__DUMMY__";
+
+
+IGameLogic::IGameLogic()
[…]

### 2018-11-05 project/blobby-1.0_fast/src/Global.h

@@ -1,0 +1,153 @@
+/*=============================================================================
+Blobby Volley 2
+Copyright (C) 2006 Jonathan Sieber (jonathan_sieber@yahoo.de)
+Copyright (C) 2006 Daniel Knobe (daniel-knobe@web.de)
+
+This program is free software; you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation; either version 2 of the License, or
+(at your option) any later version.
+
+This program is distributed in the hope that it will be useful,
+but WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+GNU General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with this program; if not, write to the Free Software
+Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
+=============================================================================*/
+
+#pragma once
+
+#include <string>
+#include <exception>
+#include <SDL2/SDL_stdinc.h>
+#include <cstring>
+
+// I hope the GP2X is the only combination of these systems
+#if defined(__linux__) && defined(__arm__)
+#define GP2X GP2X
+#endif
+
+#if defined __ANDROID__ || (defined __APPLE__ && !MAC_OS_X)
+#define __MOBILE__ true
+#define __DESKTOP__ false
+#else
+#define __MOBILE__ false
+#define __DESKTOP__ true
+#endif // defined
+
+/*!	\def DEBUG
+	\brief Enable debugging support
+	\details when this marko is present, Blobby generates some additional debugging code
+			usefull for tracking down bugs.
+*/
+
+const int BLOBBY_PORT = 1234;
+
+const int BLOBBY_VERSION_MAJOR = 0;
+const int BLOBBY_VERSION_MINOR = 103;
+
+const char AppTitle[] = "Blobby Volley 2 Version 1.0";
+const int BASE_RESOLUTION_X = 8;
+const int BASE_RESOLUTION_Y = 6;
+
+const float ROUND_START_SOUND_VOLUME = 0.2;
+const float BALL_HIT_PLAYER_SOUND_VOLUME = 0.4;
+
[…]

### 2018-11-05 project/blobby-1.0_fast/src/NetworkMessage.cpp

@@ -1,0 +1,114 @@
+/*=============================================================================
+Blobby Volley 2
+Copyright (C) 2006 Jonathan Sieber (jonathan_sieber@yahoo.de)
+Copyright (C) 2006 Daniel Knobe (daniel-knobe@web.de)
+
+This program is free software; you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation; either version 2 of the License, or
+(at your option) any later version.
+
+This program is distributed in the hope that it will be useful,
+but WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+GNU General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with this program; if not, write to the Free Software
+Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
+=============================================================================*/
+
+/* header include */
+#include "NetworkMessage.h"
+
+/* includes */
+#include <cstring>
+
+#include "UserConfig.h"
+#include "SpeedController.h"
+
+/* implementation */
+ServerInfo::ServerInfo(RakNet::BitStream& stream, const char* ip, uint16_t p)
+{
+	strncpy(hostname, ip, sizeof(hostname));
+	hostname[sizeof(hostname) - 1] = 0;
+
+	port = p;
+
+	stream.Read(activegames);
+	stream.Read(gamespeed);
+	stream.Read(name, sizeof(name));
+	stream.Read(waitingplayers);
+	stream.Read(description, sizeof(description));
+	stream.Read(rulestitle, sizeof(rulestitle));
+	stream.Read(rulesauthor, sizeof(rulesauthor));
+}
+
+ServerInfo::ServerInfo(const IUserConfigReader& config)
+{
+	/// \todo we only need a config reader here!
+
+	// default values
+	std::string n = "Blobby Volley 2 Server";
+	std::string d = "no description available";
+
+	memset(this, 0, sizeof(ServerInfo));
+	std::string tmp;
+	tmp = config.getString("name", n);
+	strncpy(name, tmp.c_str(), sizeof(name) - 1);
+	tmp = config.getString("description", d);
[…]

### 2018-11-05 project/blobby-1.0_fast/src/ScriptedInputSource.h

@@ -1,0 +1,143 @@
+/*=============================================================================
+Blobby Volley 2
+Copyright (C) 2006 Jonathan Sieber (jonathan_sieber@yahoo.de)
+Copyright (C) 2006 Daniel Knobe (daniel-knobe@web.de)
+
+This program is free software; you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation; either version 2 of the License, or
+(at your option) any later version.
+
+This program is distributed in the hope that it will be useful,
+but WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+GNU General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with this program; if not, write to the Free Software
+Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
+=============================================================================*/
+
+/**
+ * @file ScriptedInputSource.h
+ * @brief Contains a class which allows using lua scripted bots
+ */
+
+#pragma once
+
+#include <string>
+
+#include <boost/circular_buffer.hpp>
+
+#include "Global.h"
+#include "InputSource.h"
+#include "Vector.h"
+#include "IScriptableComponent.h"
+
+/// \class ScriptedInputSource
+/// \brief Bot controller
+/// \details ScriptedInputSource provides an implementation of InputSource, which uses
+/// Lua scripts to get its input. The given script is automatically initialised
+/// and provided with an interface to the game.
+
+/// The API documentation can now be found in doc/ScriptAPI.txt
+
+// The time the bot waits after game start
+const int WAITING_TIME = 00;
+
+struct lua_State;
+class DuelMatch;
+
+class ScriptedInputSource : public InputSource, public IScriptableComponent
+{
+	public:
+		/// The constructor automatically loads and initializes the script
+		/// with the given filename. The side parameter tells the script
+		/// which side is it on.
+		ScriptedInputSource(const std::string& filename, PlayerSide side, unsigned int difficulty);
+		~ScriptedInputSource();
+
+		virtual PlayerInputAbs getNextInput();
+
+	private:
+		/// this variable saves the current match
+		/// it is set each step in getInput
+		/// as in current design, only in single match mode bots are allowed
[…]

### 2018-11-05 project/blobby-1.0_fast/src/main.cpp

@@ -1,0 +1,349 @@
+/*=============================================================================
+Blobby Volley 2
+Copyright (C) 2006 Jonathan Sieber (jonathan_sieber@yahoo.de)
+Copyright (C) 2006 Daniel Knobe (daniel-knobe@web.de)
+
+This program is free software; you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation; either version 2 of the License, or
+(at your option) any later version.
+
+This program is distributed in the hope that it will be useful,
+but WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+GNU General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with this program; if not, write to the Free Software
+Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
+=============================================================================*/
+
+/* includes */
+#include <ctime>
+#include <cstring>
+#include <sstream>
+#include <iostream>
+
+#include <SDL2/SDL.h>
+
+#ifndef __APPLE__
+#ifndef __ANDROID__
+#include "config.h"
+#endif
+#endif
+
+#ifdef __APPLE__
+	#if TARGET_OS_IPHONE || TARGET_IPHONE_SIMULATOR
+		#include <physfs.h>
+	#else
+		#include "config.h"
+	#endif
+#endif
+
+#include "RenderManager.h"
+#include "SoundManager.h"
+#include "InputManager.h"
+#include "TextManager.h"
+#include "UserConfig.h"
+#include "IMGUI.h"
+#include "SpeedController.h"
+#include "Blood.h"
+#include "FileSystem.h"
+#include "state/State.h"
+
+#if defined(WIN32)
+#ifndef GAMEDATADIR
+#define GAMEDATADIR "data"
+#endif
+#endif
+
+#ifdef WIN32
+#include <windows.h>
+#undef main
+#endif
+
+/* implementation */
+
+void deinit()
+{
+	RenderManager::getSingleton().deinit();
+	SoundManager::getSingleton().deinit();
+	State::deinit();
+	SDL_Quit();
+}
+
+void setupPHYSFS()
+{
[…]

### 2018-11-05 project/blobby-1.0_fast/src/state/LocalGameState.cpp

@@ -1,0 +1,161 @@
+/*=============================================================================
+Blobby Volley 2
+Copyright (C) 2006 Jonathan Sieber (jonathan_sieber@yahoo.de)
+Copyright (C) 2006 Daniel Knobe (daniel-knobe@web.de)
+
+This program is free software; you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation; either version 2 of the License, or
+(at your option) any later version.
+
+This program is distributed in the hope that it will be useful,
+but WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+GNU General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with this program; if not, write to the Free Software
+Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
+=============================================================================*/
+
+/* header include */
+#include "LocalGameState.h"
+
+/* includes */
+#include <ctime>
+
+#include <boost/make_shared.hpp>
+
+#include "DuelMatch.h"
+#include "InputManager.h"
+#include "IMGUI.h"
+#include "ReplayRecorder.h"
+#include "SoundManager.h"
+#include "TextManager.h"
+#include "SpeedController.h"
+#include "IUserConfigReader.h"
+#include "InputSourceFactory.h"
+
+/* implementation */
+LocalGameState::~LocalGameState()
+{
+}
+
+LocalGameState::LocalGameState()
+	: mRecorder(new ReplayRecorder()), mWinner(false)
+{
+	boost::shared_ptr<IUserConfigReader> config = IUserConfigReader::createUserConfigReader("config.xml");
+	PlayerIdentity leftPlayer = config->loadPlayerIdentity(LEFT_PLAYER, false);
+	PlayerIdentity rightPlayer = config->loadPlayerIdentity(RIGHT_PLAYER, false);
+
+	boost::shared_ptr<InputSource> leftInput = InputSourceFactory::createInputSource( config, LEFT_PLAYER);
[…]

### 2018-11-05 project/blobby-1.0_fast/src/state/State.cpp

@@ -1,0 +1,218 @@
+/*=============================================================================
+Blobby Volley 2
+Copyright (C) 2006 Jonathan Sieber (jonathan_sieber@yahoo.de)
+Copyright (C) 2006 Daniel Knobe (daniel-knobe@web.de)
+
+This program is free software; you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation; either version 2 of the License, or
+(at your option) any later version.
+
+This program is distributed in the hope that it will be useful,
+but WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+GNU General Public License for more details.
+GNU General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with this program; if not, write to the Free Software
+Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA
+=============================================================================*/
+
+/* header include */
+#include "State.h"
+
+/* includes */
+#include <algorithm>
+#include <cstdio>
+
+#include <boost/make_shared.hpp>
+
+#include "LocalGameState.h"
+#include "ReplaySelectionState.h"
+#include "NetworkState.h"
+#include "NetworkSearchState.h"
+#include "OptionsState.h"
+
+#include "IMGUI.h"
+#include "TextManager.h"
+#include "SpeedController.h"
+#include "InputManager.h"
+
+
+/* implementation */
+
+// static state functions
+boost::scoped_ptr<State> State::mCurrentState(nullptr);
+boost::scoped_ptr<State> State::mStateToSwitchTo(nullptr);
+
+
+void State::deinit()
+{
+	mCurrentState.reset( nullptr );
+	mStateToSwitchTo.reset( nullptr );
+}
+
+void State::step()
+{
+	// check that we are in a valid state
+	if(mCurrentState == nullptr )
+	{
+		// otherwise, create one
[…]

### 2018-11-05 project/blobbybot.py

@@ -1,0 +1,129 @@
+import os
+import subprocess
+import numpy as np
+
+class BB:
+    def __init__(self,W,b):
+        self.W = W
+        self.b = b
+        self.fitness = np.nan
+
+
+    def l(self):
+        l = []
+        for W in self.W:
+            l.append(np.shape(W)[1])
+        l.append(np.shape(self.W[-1])[0])
+        return l
+
+
+    def compute_fitness(self):
+        self.set_bot()
+        subprocess.run('sed -i \'s/<var name="gamefps" value=".*/<var name="gamefps" value="7500"\/>/\' ~/.blobby/config.xml', shell=True)
+        os.chdir('/home/flax/blobby/blobby-1.0_fast')
+        score = subprocess.run('./src/blobby', stdout=subprocess.PIPE)
+
+        score = score.stdout.decode('utf-8').split('\n')
+        score = score[len(score)-2].split(':')
+        score_left = int(score[1])
+        score_right = int(score[2])
+        self.fitness = (score_left-score_right+25)/50
+
+        subprocess.run('sed -i \'s/<var name="gamefps" value=".*/<var name="gamefps" value="75"\/>/\' ~/.blobby/config.xml', shell=True)
+
+
+    def set_bot(self):
+        W_str = ''
+        W_str += 'W = {'
+        for i in range(len(self.W)):
+            W_str += '{'
+            for j in range(self.W[i].shape[0]):
+                W_str += '{'
+                for k in range(self.W[i].shape[1]):
+                    W_str += str(self.W[i][j,k])
+                    if k < self.W[i].shape[1]-1:
+                        W_str += ','
+                W_str += '}'
+                if j < self.W[i].shape[0]-1:
+                    W_str += ','
+            W_str += '}'
+            if i < len(self.W)-1:
+                W_str += ',\n'
+        W_str += '}\n'
+
+        b_str = ''
+        b_str += 'b = {'
+        for i in range(len(self.b)):
+            b_str += '{'
+            for j in range(self.b[i].shape[0]):
+                b_str += str(self.b[i][j][0])
+                if j < self.b[i].shape[0]-1:
+                    b_str += ','
+            b_str += '}'
+            if i < len(self.b)-1:
+                b_str += ','
+        b_str += '}\n'
+
+        f_template = open('/home/flax/blobby/nn_template.lua','r')
+        f = open('/home/flax/.blobby/scripts/nn.lua','w')
+        f.write(W_str + b_str + f_template.read())
[…]

### 2018-11-05 project/main.py

@@ -1,0 +1,105 @@
+#!/bin/python
+import numpy as np
+import matplotlib.pyplot as plt
+from blobbybot import BB
+
+l = [7,9,3] # neuronal network size
+N = 100;      # population size
+maxgen = 1000;
+sigma = 0.01;
+
+bots = [BB]*N
+offspring = [BB]*N
+fitness = np.zeros([N,maxgen])
+gen = 0
+
+for n in range(N):
+    offspring[n] = BB.random(l)
+    offspring[n].compute_fitness()
+    print(n,'/',N,', fitness: ',offspring[n].fitness)
+
+pool = offspring
+tmp = np.zeros(len(pool))
+for n in range(len(pool)):
+    tmp[n] = pool[n].fitness
+idx = np.argsort(-tmp)
+for n in range(N):
+    bots[n] = pool[idx[n]].copy()
+    fitness[n,gen] = tmp[idx[n]]
+
+
+for gen in range(1,maxgen):
+    if np.sum(fitness[:,gen-1]) == 0:
+        dist = np.arange(N)/N
+    else:
+        fbar = (fitness[:,gen-1]+0.05)**2
+        dist = np.cumsum(fbar/np.sum(fbar))
+    
+    indicator = (fitness[int(0.1*N),gen-1]-fitness[int(0.9*N),gen-1])/(fitness[0,gen-1]-fitness[N-1,gen-1])
+    print('indicator: ',indicator)
+    if gen > 4:
+        print(sigma)
+        if indicator <= 0.2:
+            sigma = sigma+0.01
+        elif indicator > 0.3:
+            sigma = sigma-0.01
+        if sigma < 0.01:
+            sigma = 0.01
+        if sigma > 0.05:
+            sigma = 0.05
+        print(sigma)
+
+
+    for n in range(int(N/2)): # crossover
+        r = np.random.rand(2);
+        for n1 in range(N):
+            if r[0] < dist[n1]:
+                break
+        for n2 in range(N):
+            if r[1] < dist[n2]:
+                break
+#        for i in range(len(bots[n].W)):
+#            offspring[2*n].W[i] = 0.2*bots[n1].W[i]+0.8*bots[n2].W[i]
+#            offspring[2*n].b[i] = 0.2*bots[n1].b[i]+0.8*bots[n2].b[i]
+#            offspring[2*n+1].W[i] = 0.8*bots[n1].W[i]+0.2*bots[n2].W[i]
[…]

### 2018-11-05 project_proposal/pp.tex
forgot some things...

@@
 \usepackage{float} % use H in figure placement
+\pagenumbering{gobble}
 
@@
 \section*{Project Proposal for Genetic Algorithms and Evolutionary Programming
-by Felix Bartel and D\` avid Kerekes}
+by Felix Bartel and D\' avid Kerekes}
 
@@
 This can for instance be given by the outcome of a game.
-Our game of choice was Blobby Volly 2 which is a simple two-dimensional vollyball game with open source written in C++ with lua support, cf. Figure \ref{fig:screenshot}.
+Our game of choice was Blobby Volley 2 which is a simple two-dimensional vollyball game with open source written in C++ with lua support, see Figure \ref{fig:screenshot}.
 
@@
 
-A scene from the game could be described by the $x$ and $y$ coordinates of the players and the ball, their corresponding velocities and the current number of ball contacts.
+A scene from the game could be described by the $x$- and $y$-coordinates of the players and the ball, their corresponding velocities and the current number of ball contacts.
 These arguments or a subset of them could then be used as input for the neuronal network.
 The output would then be the players movement which only consists of going right, left or jumping.
-Assuming the layers of the neuronal network have $n_1,\dots,n_L$ nodes we then have to optimize a function of $\sum_{l=1}^{L-1} n_ln_{l+1}$ variables if we fix the size of the network.
+Assuming the layers of the neuronal network have $n_1,\dots,n_L$ nodes we then have to optimize a function of $\sum_{l=1}^{L-1} (n_l+1)n_{l+1}$ variables if we fix the size of the network.
 A few possible extensions are
@@
 \item vary the input or output nodes
-\item vary the opponent to train (maybe also the neuronal network itself)
+\item vary the opponent or do self-training
 \end{itemize}

### 2018-11-29 project/main.py
work on

@@ -6,42 +6,37 @@ import time
 from copy import copy
 import threading
 
-l = [7,9,3] # neuronal network size
-N = 100      # population size
+l = [6,6,2] # neuronal network size
+N = 250     # population size
 maxgen = 1000
-sigma = 0.05
-np.random.seed(0)
+np.random.seed(7) # because 7 is a lucky number
+p_crossover = 0.8;
+crossover_rate = 0.5
+p_mutation = 0.03;
+mutation_rate = 0.1;
+n_elitism = 3;
 
 bots = [None]*N
 offspring = [BB.random(l, id) for id in range(N)]
 fitness = np.zeros([N,maxgen])
 
 gen_time = time.time()
-fig = plt.figure(figsize=(20,10))
+fig = plt.figure(figsize=(10,5))
 
 for gen in range(maxgen):
     if gen > 0: # no mutation or crossover for the first gen
+        print('generation time: {:1.4f}s/{:1.4f}s'.format(
+                fitness_time, time.time() - gen_time))
+        gen_time = time.time()
+
+
+# crossover using rouletta wheel
         if np.sum(fitness[:,gen-1]) == 0:
             dist = np.arange(N)/N
         else:
-            fbar = (fitness[:,gen-1]+0.05)**2
+            fbar = (fitness[:,gen-1]+0.5)**2
             dist = np.cumsum(fbar/np.sum(fbar))
 
-        indicator = (fitness[int(0.1*N),gen-1]-fitness[int(0.9*N),gen-1])/(fitness[0,gen-1]-fitness[N-1,gen-1])
-        print('generation time: {:1.4f}s/{:1.4f}s; indicator: {:0.3f}; sigma: {:0.3f}'.format(
-                fitness_time, time.time() - gen_time, indicator, sigma))
-        gen_time = time.time()
-        if gen > 4:
-            if indicator <= 0.5:
-                sigma = sigma+0.002
-            elif indicator > 0.5:
-                sigma = sigma-0.002
-            if sigma < 0.01:
-                sigma = 0.01
-            if sigma > 0.05:
-                sigma = 0.05
-
-        # crossover
         for n in range(int(N/2)):
             r = np.random.rand(2);
             for n1 in range(N):
@@ -52,25 +47,41 @@ for gen in range(maxgen):
                     break
             offspring[2*n] = BB(bots[n1].W, bots[n1].b, 2*n)
[…]

### 2018-11-30 project/bbot.py
made it pretty again and somhow got rid of bugs

@@ -1,0 +1,124 @@
+import os
+import glob
+import subprocess
+import numpy as np
+from copy import copy, deepcopy
+
+
+test = False # replace fitness function
+
+# Get a templates once at import
+with open('blobby-1.0_fast/data/.blobby/config.xml','r') as f:
+    config_template = f.read()
+
+with open('nn_template.lua','r') as f:
+    nn_template = f.read()
+
+# Clean old .xml files
+for f in glob.glob("blobby-1.0_fast/data/.blobby/config_*.xml"):
+    os.remove(f)
+
+
+def crossover_mix_nodes(bbot1, bbot2, offspring1, offspring2, rate):
+    offspring1.weights = deepcopy(bbot1.weights)
+    offspring2.weights = deepcopy(bbot2.weights)
+    for j in range(bbot1.nlayers-1):
+        for k in range(bbot1.size[j+1]):
+            if np.random.rand() < rate:
+                offspring1.weights[j][k,:] = deepcopy(bbot2.weights[j][k,:])
+                offspring2.weights[j][k,:] = deepcopy(bbot1.weights[j][k,:])
+
+
+class BBOT:
+    def __init__(self, weights, id):
+        self.weights = deepcopy(weights)
+        self.id = id
+        self.fitness = np.nan
+
+    @property
+    def id(self):
+        return self._id
+
+    @id.setter
+    def id(self, id):
+        self._id = id
+        self.config_fname = 'config_{:04d}.xml'.format(id)
+        self.config_path = 'blobby-1.0_fast/data/.blobby/' + self.config_fname
+        self.nn_path = 'blobby-1.0_fast/data/.blobby/scripts/nn_{:04d}.lua'.format(id)
+        self.config_string = config_template.replace('<var name="left_script_name" value="nn"/>',
+                                                     '<var name="left_script_name" value="nn_{:04d}"/>'.format(self.id))
+        if not os.path.isfile(self.config_path):
+            with open(self.config_path, 'w') as f:
+                f.write(self.config_string)
+
+    @property
+    def size(self):
+        # returns the size of the neuronal network
+        size = []
+        for w in self.weights:
+            size.append(np.shape(w)[1]-1)
+        size.append(np.shape(self.weights[-1])[0])
+        return size
+
+    @property
+    def nlayers(self):
+        # return the number of layers
+        return len(self.size)
+
+    @staticmethod
+    def random(size, id):
+        # creates a uniformly randon bbot
[…]

### 2018-11-30 project/bbotpool.py

@@ -1,0 +1,93 @@
+from bbot import *
+import numpy as np
+import threading
+
+
+class BBOTPOOL:
+    def __init__(self, bbots, fitness):
+        self.bbots = bbots
+        self.fitness = fitness
+
+    def __len__(self):
+        return len(self.bbots)
+
+    @staticmethod
+    def random(size, N):
+        # creates a uniformly randon bbotpool
+        bbots = [BBOT.random(size, id) for id in range(N)]
+        fitness = np.empty(N)
+        fitness.fill(np.nan)
+        return BBOTPOOL(bbots, fitness)
+
+    def __getitem__(self, key):
+        bbots = self.bbots[key]
+        fitness = self.fitness[key]
+        for n in range(len(bbots)):
+            bbots[n].id = n
+        return BBOTPOOL(bbots, fitness);
+
+    def __setitem__(self, key, pool):
+        self.bbots[key] = pool.bbots[key]
+        self.fitness[key] = pool.fitness[key]
+        for n in range(len(self.bbots)):
+            self.bbots[n].id = n
+
+    def __add__(self, pool):
+        bbots = self.bbots+pool.bbots
+        fitness = np.concatenate((self.fitness,pool.fitness))
+        for n in range(len(bbots)):
+            bbots[n].id = n
+        return BBOTPOOL(bbots, fitness);
+
+    def compute_fitness(self, mode = 0):
+        # computes the fitness of the pool
+        if mode == "all":
+            idx = range(len(self))
+        else: # don't compute fitnesses we already know
+            nofitness = np.isnan(self.fitness)
+            idx = np.nonzero(nofitness)[0]
+        threads = [threading.Thread(target=bbot.compute_fitness) for bbot in [ self.bbots[j] for j in idx]]
+        for t in threads:
+            t.start()
+        for t in threads:
+            t.join()
+        self.fitness[idx] = np.array([ bbot.fitness for bbot in [ self.bbots[j] for j in idx]])
+        self.sort()
+
+    def sort(self):
+        idx = np.argsort(-self.fitness)
+        bbots_tmp = [self.bbots[j] for j in idx]
+        fitness_tmp = self.fitness[idx]
+        self.bbots = bbots_tmp
+        self.fitness = fitness_tmp
+
+    def max2file(self):
+        self.sort()
+        self.bbots[0].set_bot('nn_max.lua')
+
+    def crossover_rouletta_wheel(self, p_crossover, rate):
+        offspring = deepcopy(self)
[…]

### 2018-11-30 project/blobbybot.py

@@ -1,131 +1,0 @@
-import os
-import glob
-import subprocess
-import numpy as np
-from copy import copy
-
-# Get a template once at import
-with open('blobby-1.0_fast/data/.blobby/config.xml','r') as f:
-    config_template = f.read()
-
-with open('nn_template.lua','r') as f:
-    nn_template = f.read()
-
-# Clean old .xml files
-for f in glob.glob("blobby-1.0_fast/data/.blobby/config_*.xml"):
-    os.remove(f)
-
-class BB:
-    def __init__(self, W, b, id):
-        self.id = id
-        self.config_fname = 'config_{:04d}.xml'.format(id)
-        self.config_path = 'blobby-1.0_fast/data/.blobby/' + self.config_fname
-        self.nn_path = 'blobby-1.0_fast/data/.blobby/scripts/nn_{:04d}.lua'.format(id)
-        self.W = copy(W)
-        self.b = copy(b)
-        self.fitness = np.nan
-
-        self.config_string = config_template.replace('<var name="left_script_name" value="nn"/>',
-                                                     '<var name="left_script_name" value="nn_{:04d}"/>'.format(self.id))
-        if not os.path.isfile(self.config_path):
-            with open(self.config_path, 'w') as f:
-                f.write(self.config_string)
-
-    def l(self):
-        l = []
-        for W in self.W:
-            l.append(np.shape(W)[1])
-        l.append(np.shape(self.W[-1])[0])
-        return l
-
-    def compute_fitness(self):
-        self.set_bot()
-
-        score = subprocess.run(['./src/blobby', self.config_fname],
-                               stdout=subprocess.PIPE,
-                               cwd='blobby-1.0_fast')
-
-        score = score.stdout.decode('utf-8').split('\n')
-        score = score[len(score)-2].split(':')
-        score_left = int(score[1])
-        score_right = int(score[2])
-        self.fitness = score_left/(score_left+score_right)
-
-    def set_bot(self):
-        W_str = ''
-        W_str += 'W = {'
-        for i in range(len(self.W)):
-            W_str += '{'
-            for j in range(self.W[i].shape[0]):
-                W_str += '{'
-                for k in range(self.W[i].shape[1]):
-                    W_str += str(self.W[i][j,k])
-                    if k < self.W[i].shape[1]-1:
-                        W_str += ','
-                W_str += '}'
-                if j < self.W[i].shape[0]-1:
-                    W_str += ','
[…]

### 2018-11-30 project/little_functions.py

@@ -1,0 +1,24 @@
+import matplotlib.pyplot as plt
+import numpy as np
+
+
+def plot(fig, fitness):
+    gen = len(fitness)
+    if gen == 1:
+        return
+    fig.clf()
+
+    ax = plt.axes()
+    ax.plot(np.arange(gen),np.mean(fitness,axis=1),'k')
+    for x in [0,0.1,0.25]:
+        top = np.array([f[int(x*len(f))] for f in fitness])
+        bot = np.array([f[int((1-x)*len(f)-1)] for f in fitness])
+        ax.fill_between(np.arange(gen),bot-0.01,top+0.01,facecolor='blue',alpha=0.2)
+
+    ax.set_xlabel('generation')
+    ax.set_ylabel('fitness')
+    ax.set_xlim([0,gen-1])
+    ax.set_ylim([0,1])
+
+    plt.show(block=False)
+    plt.pause(0.1)

### 2018-11-30 project/main.py

@@ -1,115 +1,42 @@
-#!/bin/python
+#!/bin/bash
+from bbotpool import *
+from little_functions import *
 import numpy as np
 import matplotlib.pyplot as plt
-from blobbybot import BB
 import time
-from copy import copy
-import threading
 
-l = [6,6,2] # neuronal network size
-N = 250     # population size
-maxgen = 1000
-np.random.seed(7) # because 7 is a lucky number
-p_crossover = 0.8;
-crossover_rate = 0.5
-p_mutation = 0.03;
-mutation_rate = 0.1;
-n_elitism = 3;
 
-bots = [None]*N
-offspring = [BB.random(l, id) for id in range(N)]
-fitness = np.zeros([N,maxgen])
+np.random.seed(7)       # because 7 is a lucky number
+size = [6,6,2]        # size of the neuronal networks
+N = 100                 # population size
+maxgen = 100            # maximal number of generation
+p_crossover = 0.9       # crossove parameters
+crossover_rate = 0.5    
+p_mutation = 0.7
+mutation_rate = 0.05    #mutation parameters
+mutation_sigma = 0.1    
+n_elitism = 5           # keep the best individuals
 
-gen_time = time.time()
 fig = plt.figure(figsize=(10,5))
+pool = BBOTPOOL.random(size, N)
+pool.compute_fitness()
+fitness = [pool.fitness]
+
 
 for gen in range(maxgen):
-    if gen > 0: # no mutation or crossover for the first gen
-        print('generation time: {:1.4f}s/{:1.4f}s'.format(
-                fitness_time, time.time() - gen_time))
-        gen_time = time.time()
-
-
-# crossover using rouletta wheel
-        if np.sum(fitness[:,gen-1]) == 0:
-            dist = np.arange(N)/N
-        else:
-            fbar = (fitness[:,gen-1]+0.5)**2
-            dist = np.cumsum(fbar/np.sum(fbar))
-
-        for n in range(int(N/2)):
-            r = np.random.rand(2);
-            for n1 in range(N):
-                if r[0] < dist[n1]:
-                    break
-            for n2 in range(N):
-                if r[1] < dist[n2]:
-                    break
-            offspring[2*n] = BB(bots[n1].W, bots[n1].b, 2*n)
-            offspring[2*n+1] = BB(bots[n2].W, bots[n2].b, 2*n+1)
-            if np.random.rand() < p_crossover:
[…]

### 2018-11-30 project/main.py
minor

@@ -7,14 +7,14 @@ import time
 
 
 np.random.seed(7)       # because 7 is a lucky number
-size = [6,6,2]        # size of the neuronal networks
+size = [6,7,2]          # size of the neuronal networks
 N = 100                 # population size
 maxgen = 100            # maximal number of generation
 p_crossover = 0.9       # crossove parameters
-crossover_rate = 0.5    
+crossover_rate = 0.3
 p_mutation = 0.7
-mutation_rate = 0.05    #mutation parameters
-mutation_sigma = 0.1    
+mutation_rate = 0.05    # mutation parameters
+mutation_sigma = 0.05
 n_elitism = 5           # keep the best individuals
 
 fig = plt.figure(figsize=(10,5))

### 2018-11-30 project/bbotpool.py
bug fixes, tweaks, huge improvements :-)

@@ -19,6 +19,14 @@ class BBOTPOOL:
         fitness.fill(np.nan)
         return BBOTPOOL(bbots, fitness)
 
+    @staticmethod
+    def zeros(N):
+        # creates an empty bbotpool
+        bbots = [BBOT(0, id) for id in range(N)]
+        fitness = np.empty(N)
+        fitness.fill(np.nan)
+        return BBOTPOOL(bbots, fitness)
+
     def __getitem__(self, key):
         bbots = self.bbots[key]
         fitness = self.fitness[key]
@@ -27,8 +35,8 @@ class BBOTPOOL:
         return BBOTPOOL(bbots, fitness);
 
     def __setitem__(self, key, pool):
-        self.bbots[key] = pool.bbots[key]
-        self.fitness[key] = pool.fitness[key]
+        self.bbots[key] = pool.bbots
+        self.fitness[key] = pool.fitness
         for n in range(len(self.bbots)):
             self.bbots[n].id = n
 
@@ -46,44 +54,57 @@ class BBOTPOOL:
         else: # don't compute fitnesses we already know
             nofitness = np.isnan(self.fitness)
             idx = np.nonzero(nofitness)[0]
-        threads = [threading.Thread(target=bbot.compute_fitness) for bbot in [ self.bbots[j] for j in idx]]
-        for t in threads:
-            t.start()
-        for t in threads:
-            t.join()
-        self.fitness[idx] = np.array([ bbot.fitness for bbot in [ self.bbots[j] for j in idx]])
+        chunks = list(range(0, len(idx), 50)) # maximal parallel threads
+        chunks.append(len(idx))
+
+        for j in range(len(chunks)-1):
+            idx_chunk = [ idx[n] for n in range(chunks[j],chunks[j+1]) ]
+            threads = [threading.Thread(target=bbot.compute_fitness) for bbot in [ self.bbots[j] for j in idx_chunk]]
+            for t in threads:
+                t.start()
+            for t in threads:
+                t.join()
+            self.fitness[idx_chunk] = np.array([ bbot.fitness for bbot in [ self.bbots[j] for j in idx_chunk]])
         self.sort()
 
     def sort(self):
         idx = np.argsort(-self.fitness)
-        bbots_tmp = [self.bbots[j] for j in idx]
-        fitness_tmp = self.fitness[idx]
-        self.bbots = bbots_tmp
-        self.fitness = fitness_tmp
+        bbots = [self.bbots[j] for j in idx]
[…]

### 2018-11-30 project/little_functions.py

@@ -10,10 +10,10 @@ def plot(fig, fitness):
 
     ax = plt.axes()
     ax.plot(np.arange(gen),np.mean(fitness,axis=1),'k')
-    for x in [0,0.1,0.25]:
+    for x in np.arange(0,0.25,0.01):
         top = np.array([f[int(x*len(f))] for f in fitness])
         bot = np.array([f[int((1-x)*len(f)-1)] for f in fitness])
-        ax.fill_between(np.arange(gen),bot-0.01,top+0.01,facecolor='blue',alpha=0.2)
+        ax.fill_between(np.arange(gen),bot-0.01,top+0.01,facecolor='blue',alpha=0.04)
 
     ax.set_xlabel('generation')
     ax.set_ylabel('fitness')

### 2018-11-30 project/main.py

@@ -7,36 +7,40 @@ import time
 
 
 np.random.seed(7)       # because 7 is a lucky number
-size = [6,7,2]          # size of the neuronal networks
+size = [6, 7, 2]        # size of the neuronal networks
 N = 100                 # population size
 maxgen = 100            # maximal number of generation
+N_offspring = 150
 p_crossover = 0.9       # crossove parameters
 crossover_rate = 0.3
-p_mutation = 0.7
-mutation_rate = 0.05    # mutation parameters
-mutation_sigma = 0.05
-n_elitism = 5           # keep the best individuals
+p_mutation = 0.5
+mutation_rate = 0.5     # mutation parameters
+mutation_sigma = 0.01
+n_elitism = 30          # keep the best individuals
 
 fig = plt.figure(figsize=(10,5))
 pool = BBOTPOOL.random(size, N)
-pool.compute_fitness()
+pool.compute_fitness('all')
 fitness = [pool.fitness]
 
 
 for gen in range(maxgen):
     gen_time = time.time()
-    offspring = pool.crossover_rouletta_wheel(p_crossover, crossover_rate)
+    offspring = pool.crossover_rouletta_wheel(N_offspring, p_crossover, crossover_rate)
     offspring.mutate_gaussian(p_mutation, mutation_rate, mutation_sigma)
     pool = deepcopy(pool[0:n_elitism]+offspring)
     fitness_time = time.time()
-    pool.compute_fitness('all')
+    if gen%5 == 0:
+        pool.compute_fitness('all')
+    else:
+        pool.compute_fitness()
     fitness_time = time.time() - fitness_time
     pool = pool[0:N]
-    pool.max2file()
-    fitness.append(list(pool.fitness))
+    fitness.append(pool.fitness)
 
     print('generation time: {:1.4f}s/{:1.4f}s'.format(
             fitness_time, time.time() - gen_time))
     plot(fig, fitness)
+    pool.max2file()
 
 plt.show()

### 2018-12-01 project/bbotpool.py
modified plotting

@@ -83,7 +83,7 @@ class BBOTPOOL:
         if np.sum(self.fitness) == 0:
             dist = np.arange(len(self))/len(self) # uniform wheel
         else:
-            fbar = (self.fitness+1)**2
+            fbar = (self.fitness+1)
             dist = np.cumsum(fbar/np.sum(fbar))
 
         idx = []

### 2018-12-01 project/little_functions.py

@@ -9,15 +9,24 @@ def plot(fig, fitness):
     fig.clf()
 
     ax = plt.axes()
-    ax.plot(np.arange(gen),np.mean(fitness,axis=1),'k')
-    for x in np.arange(0,0.25,0.01):
-        top = np.array([f[int(x*len(f))] for f in fitness])
-        bot = np.array([f[int((1-x)*len(f)-1)] for f in fitness])
-        ax.fill_between(np.arange(gen),bot-0.01,top+0.01,facecolor='blue',alpha=0.04)
+    ax.plot(np.arange(gen), np.mean(fitness, axis = 1), 'k')
+    ax.plot(np.arange(gen), np.max(fitness, axis = 1), 'k')
+    ax.plot(np.arange(gen), np.min(fitness, axis = 1), 'k')
+
+    yres = 20
+    y_grid = np.linspace(0, 1, yres)
+    zi = [ [ np.sum(np.less_equal(y_grid[j],fitness[k])*np.less(fitness[k],y_grid[j+1])) for j in range(len(y_grid)-1)] for k in range(gen) ]
+
+    zi = np.transpose(np.array(zi))
+    zi = np.concatenate((zi, np.zeros([1,gen])))
+    xi, yi = np.meshgrid(np.arange(0,gen)-0.5, y_grid)
+    tmp = ax.pcolor(xi, yi, zi, cmap=plt.cm.Oranges)
+    plt.colorbar(tmp)
+    tmp.set_clim(vmin=0, vmax=20)
 
     ax.set_xlabel('generation')
     ax.set_ylabel('fitness')
-    ax.set_xlim([0,gen-1])
+    ax.set_xlim([0,gen-1.5])
     ax.set_ylim([0,1])
 
     plt.show(block=False)

### 2018-12-01 project/main.py

@@ -12,7 +12,7 @@ N = 100                 # population size
 maxgen = 100            # maximal number of generation
 N_offspring = 150
 p_crossover = 0.9       # crossove parameters
-crossover_rate = 0.3
+crossover_rate = 0.5
 p_mutation = 0.5
 mutation_rate = 0.5     # mutation parameters
 mutation_sigma = 0.01

### 2018-12-01 project/main.py
minor

@@ -16,7 +16,7 @@ crossover_rate = 0.5
 p_mutation = 0.5
 mutation_rate = 0.5     # mutation parameters
 mutation_sigma = 0.01
-n_elitism = 30          # keep the best individuals
+n_elitism = 10          # keep the best individuals
 
 fig = plt.figure(figsize=(10,5))
 pool = BBOTPOOL.random(size, N)

### 2018-12-01 project_proposal/pp.tex
updated project proposal

@@
 
-Our idea was to apply a genetic algorithm to a neuronal network which is known as neuroevolution.
-A big advantage of this sheme is that one only needs an indicator how good the neuronal network is given a certain task.
+We want to apply a genetic algorithm to a neuronal network which is known as neuroevolution.
+A big advantage of this scheme is that one only needs an indicator how good the neuronal network is given a certain task.
 This can for instance be given by the outcome of a game.
-Our game of choice was Blobby Volley 2 which is a simple two-dimensional vollyball game with open source written in C++ with lua support, see Figure \ref{fig:screenshot}.
+Our game of choice was Blobby Volley 2 which is a simple two-dimensional volleyball game with open source written in C++ with lua support, see Figure \ref{fig:screenshot} (a).
 
@@
 \center
-\includegraphics[width=0.7\textwidth]{img/screenshot.png}
-\caption{screenshot of the game}
+\begin{tabular}{cc}
+\includegraphics[width=0.3\textwidth]{img/screenshot.png}
+&
+\includegraphics[width=0.4\textwidth]{img/fitness.png}
+\\
+(a) & (b)
+\end{tabular}
+\caption{(a) screenshot of the game; (b) fitness density of an example run with indicated minimum, average and maximum}
 \label{fig:screenshot}
@@
 Assuming the layers of the neuronal network have $n_1,\dots,n_L$ nodes we then have to optimize a function of $\sum_{l=1}^{L-1} (n_l+1)n_{l+1}$ variables if we fix the size of the network.
-A few possible extensions are
-\begin{itemize}
-\item different crossover operators
-\item differnet mutation operators (maybe use a covanriance matrix, if the size allows it)
-\item make the size of the network variable
-\item vary the input or output nodes
-\item vary the opponent or do self-training
-\end{itemize}
+
+We have done some tests so far with a basic evolutionary algorithm where we fixed the size of the neuronal network to $6$ input nodes, one hidden layer with $7$ nodes and $2$ output nodes.
+As crossover we randomly mixed the weights and biases of two networks and chose the parents via the fitness proportional roulette wheel selection.
+Mutation is accomplished by adding Gaussian noise to a subset of the nodes.
+Furthermore we implemented elitism.
+
+Because a random neuronal network has barely a change against the given bots in the game, we implemented our own trainer.
+The trainer is easier to defeat and even gives up after a certain amount of touches with the ball such that normal gameplay will be rewarded.
+This simplification gives the evolutionary algorithm the opportunity to have at least some success rate on which it can hook on.
+The results are depicted in Figure \ref{fig:screenshot} (b), where the fitness is the normalized point difference after $21$ serves.
+
+One can see that, although we we made huge simplifications by weakening the opponent, the algorithm produces a network which wins all balls.
+
+We could extend this work by e.g.\,making the size of the network variable or do self-training (maybe with similar techniques as in the paper of Karl Sims which was presented in the lecture)
 

### 2018-12-03 project_proposal/pp.tex
Add refs, update proposal.

--- 2018-11-05 project_proposal/pp.tex
@@
 %{{{ general stuff
+\usepackage[a4paper,bindingoffset=0.2in,%
+            left=1in,right=1in,top=1in,bottom=1in,%
+            footskip=.25in]{geometry}
 \usepackage[english]{babel}
@@
 \usepackage[T1]{fontenc}
+\usepackage{cite}
 \usepackage[hidelinks]{hyperref}
@@
 \date{}
+\title{Project Proposal for Genetic Algorithms and Evolutionary Programming}
+\author{Felix Bartel and D\'avid Kerekes}
 
@@
 
-\section*{Project Proposal for Genetic Algorithms and Evolutionary Programming
-by Felix Bartel and D\' avid Kerekes}
+\maketitle
 
-Our idea was to apply a genetic algorithm to a neuronal network which is known as neuroevolution.
-A big advantage of this sheme is that one only needs an indicator how good the neuronal network is given a certain task.
+\section{Problem Overview}
+
+We want to apply a genetic algorithm to a neural network which is known as neuroevolution.
+A big advantage of this scheme is that one only needs an indicator how good the neural network is given a certain task.
 This can for instance be given by the outcome of a game.
-Our game of choice was Blobby Volley 2 which is a simple two-dimensional vollyball game with open source written in C++ with lua support, see Figure \ref{fig:screenshot}.
+Our game of choice was Blobby Volley 2 which is a simple two-dimensional volleyball game with open source written in C++ with lua support, see Figure \ref{fig:screenshot}.
 
@@
 \includegraphics[width=0.7\textwidth]{img/screenshot.png}
-\caption{screenshot of the game}
+\caption{Screenshot of the game.}
 \label{fig:screenshot}
@@
 A scene from the game could be described by the $x$- and $y$-coordinates of the players and the ball, their corresponding velocities and the current number of ball contacts.
-These arguments or a subset of them could then be used as input for the neuronal network.
+These arguments or a subset of them could then be used as input for the neural network.
 The output would then be the players movement which only consists of going right, left or jumping.
-Assuming the layers of the neuronal network have $n_1,\dots,n_L$ nodes we then have to optimize a function of $\sum_{l=1}^{L-1} (n_l+1)n_{l+1}$ variables if we fix the size of the network.
-A few possible extensions are
-\begin{itemize}
-\item different crossover operators
-\item differnet mutation operators (maybe use a covanriance matrix, if the size allows it)
-\item make the size of the network variable
-\item vary the input or output nodes
-\item vary the opponent or do self-training
-\end{itemize}
+Assuming the layers of the neural network have $n_1,\dots,n_L$ nodes we then have to optimize \[N = \sum_{l=1}^{L-1} (n_l+1) \, n_{l+1}\] number of variables if we fix the size of the network.
+
+\section*{Preliminary Results}
+
+We have done some tests so far with a basic evolutionary algorithm \cite{github_repo} where we fixed the size of the neural network to $6$ input nodes, one hidden layer with $7$ nodes and $2$ output nodes.
+As crossover we randomly mixed the weights and biases of two networks and chose the parents via the fitness proportional roulette wheel selection.
+Mutation is accomplished by adding Gaussian noise to a subset of the nodes.
+Furthermore we implemented elitism.
+
+The representation of the network is accomplished through python objects.
+This will allow a straightforward implementation of a wide variety of genetic operators, while having next to no impact on the speed of training.
+Evaluation of the population is done in a thread-parallel fashion, allowing us to exploit multiple high core-count machines at our disposal.
+
+We implemented our own very bot to act as an initial trainer for the network.
+This trainer is easier to defeat and even gives up after a certain amount of touches with the ball such that normal gameplay will be rewarded.
+The simplification gives the evolutionary algorithm the opportunity to start optimizing the neural network in a valid direction, without introducing additional factors to the fitness score.
+
+\begin{figure}[H]
+\center
+\includegraphics[width=0.9\textwidth]{img/fitness.png}
+\caption{Fitness density of an example run with indicated minimum, average and maximum.}
+\label{fig:train}
+\end{figure}
+
+The results are depicted in Figure \ref{fig:train}, where the fitness is the normalized point difference after $21$ serves.
+One can see that, although we we made simplifications by weakening the opponent, the algorithm produces a network which wins all rounds.
+
+\section*{Further Experiments}
+
+We plan on testing multiple types of crossover and mutation operators, and selection schemes as seen in \cite{montana1989training}.
+Examples mutations include biased and unbiased weight mutations and node mutations.
+Example crossovers are weight, node, and feature crossovers.
+
+To have a more objective representation of the capabilities of the networks, we plan on reintroducing the original easy, medium and hard difficulty bots in a curriculum-learning fashion.
+
+Finally, as time allows we could extend this work by e.g.\,making the size of the network variable or do self-training with similar techniques as in the paper of Karl Sims\cite{sims1994evolving} presented in the lectures.
+
+\bibliography{refs}{}
+\bibliographystyle{plain}
 

### 2018-12-04 project/bbot.py
Fix unpythonic object naming scheme

@@ -29,7 +29,7 @@ def crossover_mix_nodes(bbot1, bbot2, offspring1, offspring2, rate):
                 offspring2.weights[j][k,:] = deepcopy(bbot1.weights[j][k,:])
 
 
-class BBOT:
+class Bot:
     def __init__(self, weights, id):
         self.weights = deepcopy(weights)
         self.id = id
@@ -71,7 +71,7 @@ class BBOT:
         weights = []
         for j in range(len(size)-1):
             weights.append(np.random.rand(size[j+1], size[j]+1)-0.5)
-        return BBOT(weights, id)
+        return Bot(weights, id)
 
     def mutate_gaussian(self, rate, sigma):
         # adds Gaussian noise with deviation sigma to rate weights (0.5 for half of them)

### 2018-12-04 project/bbotpool.py

@@ -3,7 +3,7 @@ import numpy as np
 import threading
 
 
-class BBOTPOOL:
+class BotPool:
     def __init__(self, bbots, fitness):
         self.bbots = bbots
         self.fitness = fitness
@@ -14,25 +14,25 @@ class BBOTPOOL:
     @staticmethod
     def random(size, N):
         # creates a uniformly randon bbotpool
-        bbots = [BBOT.random(size, id) for id in range(N)]
+        bbots = [Bot.random(size, id) for id in range(N)]
         fitness = np.empty(N)
         fitness.fill(np.nan)
-        return BBOTPOOL(bbots, fitness)
+        return BotPool(bbots, fitness)
 
     @staticmethod
     def zeros(N):
         # creates an empty bbotpool
-        bbots = [BBOT(0, id) for id in range(N)]
+        bbots = [Bot(0, id) for id in range(N)]
         fitness = np.empty(N)
         fitness.fill(np.nan)
-        return BBOTPOOL(bbots, fitness)
+        return BotPool(bbots, fitness)
 
     def __getitem__(self, key):
         bbots = self.bbots[key]
         fitness = self.fitness[key]
         for n in range(len(bbots)):
             bbots[n].id = n
-        return BBOTPOOL(bbots, fitness);
+        return BotPool(bbots, fitness);
 
     def __setitem__(self, key, pool):
         self.bbots[key] = pool.bbots
@@ -45,7 +45,7 @@ class BBOTPOOL:
         fitness = np.concatenate((self.fitness,pool.fitness))
         for n in range(len(bbots)):
             bbots[n].id = n
-        return BBOTPOOL(bbots, fitness);
+        return BotPool(bbots, fitness);
 
     def compute_fitness(self, mode = 0):
         # computes the fitness of the pool
@@ -93,7 +93,7 @@ class BBOTPOOL:
                     idx.append(n)
                     break
 
-        offspring = BBOTPOOL.zeros(N)
+        offspring = BotPool.zeros(N)
         for n in range(int(N/2)):
             if np.random.rand() < p_crossover:
                 crossover_mix_nodes(self.bbots[idx[2*n]], self.bbots[idx[2*n+1]], offspring.bbots[2*n], offspring.bbots[2*n+1], rate)

### 2018-12-04 project/main.py

@@ -19,7 +19,7 @@ mutation_sigma = 0.01
 n_elitism = 10          # keep the best individuals
 
 fig = plt.figure(figsize=(10,5))
-pool = BBOTPOOL.random(size, N)
+pool = BotPool.random(size, N)
 pool.compute_fitness('all')
 fitness = [pool.fitness]
 

### 2018-12-04 project/bbot.py
Clean up lua writing function

@@ -7,7 +7,7 @@ from copy import copy, deepcopy
 
 test = False # replace fitness function
 
-# Get a templates once at import
+# Get templates once at import
 with open('blobby-1.0_fast/data/.blobby/config.xml','r') as f:
     config_template = f.read()
 
@@ -87,7 +87,7 @@ class Bot:
                 frob += np.sum(w**2)
             self.fitness = 1/(1+frob)
         else:
-            self.set_bot()
+            self.write_lua()
 
             score = subprocess.run(['./src/blobby', self.config_fname],
                                    stdout=subprocess.PIPE,
@@ -99,26 +99,25 @@ class Bot:
             score_right = int(score[2])
             self.fitness = score_left/(score_left+score_right)
 
-    def set_bot(self, path = 0):
-        if path == 0:
+    def write_lua(self, path = None):
+        if not path: # Works for empty sting, empty list...
             path = self.nn_path
-        W_str = ''
-        W_str += 'W = {'
-        for i in range(self.nlayers-1):
+        W_str = 'W = {'
+        for i in range(self.nlayers - 1):
             W_str += '{'
             for j in range(self.size[i+1]):
                 W_str += '{'
                 for k in range(self.size[i]):
-                    W_str += str(self.weights[i][j,k])
+                    W_str += str(self.weights[i][j, k])
                     if k < self.size[i]-1:
                         W_str += ','
                 W_str += '}'
                 if j < self.size[i+1]:
                     W_str += ','
             W_str += '}'
-            if i < self.nlayers-1:
+            if i < self.nlayers - 1:
                 W_str += ',\n'
         W_str += '}\n'
 
-        with open(path,'w') as f:
+        with open(path, 'w') as f:
             f.write(W_str + nn_template)

### 2018-12-04 project/bbotpool.py

@@ -69,14 +69,12 @@ class BotPool:
 
     def sort(self):
         idx = np.argsort(-self.fitness)
-        bbots = [self.bbots[j] for j in idx]
-        fitness = self.fitness[idx]
-        self.bbots = bbots
-        self.fitness = fitness
+        self.bbots = [self.bbots[i] for i in idx]
+        self.fitness = self.fitness[idx]
 
     def max2file(self):
         self.sort()
-        self.bbots[0].set_bot('nn_max.lua')
+        self.bbots[0].write_lua('nn_max.lua')
 
     def crossover_rouletta_wheel(self, N, p_crossover, rate):
         N = round(N/2)*2

### 2018-12-04 project/bot.py
Clean up __getitem__ of botpool.py, rename pys to keep up with object naming.

@@ -18,17 +18,6 @@ with open('nn_template.lua','r') as f:
 for f in glob.glob("blobby-1.0_fast/data/.blobby/config_*.xml"):
     os.remove(f)
 
-
-def crossover_mix_nodes(bbot1, bbot2, offspring1, offspring2, rate):
-    offspring1.weights = deepcopy(bbot1.weights)
-    offspring2.weights = deepcopy(bbot2.weights)
-    for j in range(bbot1.nlayers-1):
-        for k in range(bbot1.size[j+1]):
-            if np.random.rand() < rate:
-                offspring1.weights[j][k,:] = deepcopy(bbot2.weights[j][k,:])
-                offspring2.weights[j][k,:] = deepcopy(bbot1.weights[j][k,:])
-
-
 class Bot:
     def __init__(self, weights, id):
         self.weights = deepcopy(weights)

### 2018-12-04 project/botpool.py

@@ -6,6 +6,7 @@ import threading
 class BotPool:
     def __init__(self, bbots, fitness):
         self.bbots = bbots
+        # TODO Why do we keeps fitnesses seperately here and in the class?
         self.fitness = fitness
 
     def __len__(self):
@@ -13,7 +14,7 @@ class BotPool:
 
     @staticmethod
     def random(size, N):
-        # creates a uniformly randon bbotpool
+        # creates a uniformly random bbotpool
         bbots = [Bot.random(size, id) for id in range(N)]
         fitness = np.empty(N)
         fitness.fill(np.nan)
@@ -30,9 +31,12 @@ class BotPool:
     def __getitem__(self, key):
         bbots = self.bbots[key]
         fitness = self.fitness[key]
-        for n in range(len(bbots)):
-            bbots[n].id = n
-        return BotPool(bbots, fitness);
+        if hasattr(bbots, "__iter__"):  # Check if we got a list or a single object
+            for n in range(len(bbots)):
+                bbots[n].id = n
+            return BotPool(bbots, fitness)
+        else:
+            return bbots, fitness
 
     def __setitem__(self, key, pool):
         self.bbots[key] = pool.bbots
@@ -45,7 +49,7 @@ class BotPool:
         fitness = np.concatenate((self.fitness,pool.fitness))
         for n in range(len(bbots)):
             bbots[n].id = n
-        return BotPool(bbots, fitness);
+        return BotPool(bbots, fitness)
 
     def compute_fitness(self, mode = 0):
         # computes the fitness of the pool

### 2018-12-04 project/main.py

@@ -1,9 +1,10 @@
 #!/bin/bash
-from bbotpool import *
-from little_functions import *
+from botpool import BotPool
+from little_functions import plot
 import numpy as np
 import matplotlib.pyplot as plt
 import time
+from copy import deepcopy
 
 
 np.random.seed(7)       # because 7 is a lucky number
@@ -38,7 +39,8 @@ for gen in range(maxgen):
     pool = pool[0:N]
     fitness.append(pool.fitness)
 
-    print('generation time: {:1.4f}s/{:1.4f}s'.format(
+    print('Gen {}/{}; time: {:1.4f}s/{:1.4f}s'.format(
+            gen, maxgen,
             fitness_time, time.time() - gen_time))
     plot(fig, fitness)
     pool.max2file()

### 2018-12-04 project/botpool.py
Fix by adding second half of previous commit

@@ -1,7 +1,16 @@
-from bbot import *
+from bot import Bot
 import numpy as np
 import threading
+from copy import deepcopy
 
+def crossover_mix_nodes(bbot1, bbot2, offspring1, offspring2, rate):
+    offspring1.weights = deepcopy(bbot1.weights)
+    offspring2.weights = deepcopy(bbot2.weights)
+    for j in range(bbot1.nlayers-1):
+        for k in range(bbot1.size[j+1]):
+            if np.random.rand() < rate:
+                offspring1.weights[j][k,:] = deepcopy(bbot2.weights[j][k,:])
+                offspring2.weights[j][k,:] = deepcopy(bbot1.weights[j][k,:])
 
 class BotPool:
     def __init__(self, bbots, fitness):

### 2018-12-04 project/botpool.py
Btw it's roulette

@@ -89,7 +89,7 @@ class BotPool:
         self.sort()
         self.bbots[0].write_lua('nn_max.lua')
 
-    def crossover_rouletta_wheel(self, N, p_crossover, rate):
+    def crossover_roulette_wheel(self, N, p_crossover, rate):
         N = round(N/2)*2
         if np.sum(self.fitness) == 0:
             dist = np.arange(len(self))/len(self) # uniform wheel

### 2018-12-04 project/main.py

@@ -27,7 +27,7 @@ fitness = [pool.fitness]
 
 for gen in range(maxgen):
     gen_time = time.time()
-    offspring = pool.crossover_rouletta_wheel(N_offspring, p_crossover, crossover_rate)
+    offspring = pool.crossover_roulette_wheel(N_offspring, p_crossover, crossover_rate)
     offspring.mutate_gaussian(p_mutation, mutation_rate, mutation_sigma)
     pool = deepcopy(pool[0:n_elitism]+offspring)
     fitness_time = time.time()

### 2018-12-04 project_proposal/pp.tex
Fix a typo in proposal
minors in the proposal

@@
 We want to apply a genetic algorithm to a neural network which is known as neuroevolution.
-A big advantage of this scheme is that one only needs an indicator how good the neural network is given a certain task.
+A big advantage of this scheme is that one only needs an indicator how good the neural network performs given a certain task.
 This can for instance be given by the outcome of a game.
@@
 \center
-\includegraphics[width=0.7\textwidth]{img/screenshot.png}
+\includegraphics[width=0.4\textwidth]{img/screenshot.png}
 \caption{Screenshot of the game.}
@@
 We have done some tests so far with a basic evolutionary algorithm \cite{github_repo} where we fixed the size of the neural network to $6$ input nodes, one hidden layer with $7$ nodes and $2$ output nodes.
-As crossover we randomly mixed the weights and biases of two networks and chose the parents via the fitness proportional roulette wheel selection.
-Mutation is accomplished by adding Gaussian noise to a subset of the nodes.
+As crossover we randomly mixed the nodes with belonging weights and biases of two networks and chose the parents via the fitness proportional roulette wheel selection.
+Mutation is accomplished by adding Gaussian noise to a subset of all weights and biases.
 Furthermore we implemented elitism.
 
-The representation of the network is accomplished through python objects.
+The genotype-representation of the network is accomplished through python objects.
 This will allow a straightforward implementation of a wide variety of genetic operators, while having next to no impact on the speed of training.
-Evaluation of the population is done in a thread-parallel fashion, allowing us to exploit multiple high core-count machines at our disposal.
+As the evaluation of the population is time-consuming, we do this in a thread-parallel fashion, allowing us to exploit multiple high core-count machines at our disposal.
 
-We implemented our own very bot to act as an initial trainer for the network.
+We implemented our own bot to act as an initial trainer for the network.
 This trainer is easier to defeat and even gives up after a certain amount of touches with the ball such that normal gameplay will be rewarded.
@@
 \center
-\includegraphics[width=0.9\textwidth]{img/fitness.png}
+\includegraphics[width=0.5\textwidth]{img/fitness.png}
 \caption{Fitness density of an example run with indicated minimum, average and maximum.}
@@
 We plan on testing multiple types of crossover and mutation operators, and selection schemes as seen in \cite{montana1989training}.
-Examples mutations include biased and unbiased weight mutations and node mutations.
+Example mutations include biased and unbiased weight mutations and node mutations.
 Example crossovers are weight, node, and feature crossovers.
@@
 
-Finally, as time allows we could extend this work by e.g.\,making the size of the network variable or do self-training with similar techniques as in the paper of Karl Sims\cite{sims1994evolving} presented in the lectures.
+Finally, as time allows we could extend this work by e.g.\,making the size of the network variable or do self-training with similar techniques as in the paper of Karl Sims\cite{sims1994evolving} presented in the lecture.
 

### 2018-12-04 project_proposal/pp.tex
Small cosmetic fixes

--- 2018-12-03 project_proposal/pp.tex
@@
 
-We implemented our own very bot to act as an initial trainer for the network.
-This trainer is easier to defeat and even gives up after a certain amount of touches with the ball such that normal gameplay will be rewarded.
+We implemented our own bot to act as an initial trainer for the network.
+This trainer is easier to defeat and even gives up after a certain amount of touches with the ball, such that normal gameplay will be rewarded.
 The simplification gives the evolutionary algorithm the opportunity to start optimizing the neural network in a valid direction, without introducing additional factors to the fitness score.

### 2018-12-04 project/botpool.py

@@ -60,7 +60,7 @@ class BotPool:
             bbots[n].id = n
         return BotPool(bbots, fitness)
 
-    def compute_fitness(self, mode = 0):
+    def compute_fitness(self, mode=None):
         # computes the fitness of the pool
         if mode == "all":
             idx = range(len(self))

### 2018-12-04 project/bot.py
Rework fitness keeping

@@ -22,7 +22,7 @@ class Bot:
     def __init__(self, weights, id):
         self.weights = deepcopy(weights)
         self.id = id
-        self.fitness = np.nan
+        self.fitness = -np.inf
 
     @property
     def id(self):
@@ -63,6 +63,7 @@ class Bot:
         return Bot(weights, id)
 
     def mutate_gaussian(self, rate, sigma):
+        self.fitness = -np.inf
         # adds Gaussian noise with deviation sigma to rate weights (0.5 for half of them)
         for j in range(self.nlayers-1):
             mask = np.less(np.random.random_sample(self.weights[j].shape), rate)

### 2018-12-04 project/botpool.py

@@ -13,10 +13,8 @@ def crossover_mix_nodes(bbot1, bbot2, offspring1, offspring2, rate):
                 offspring2.weights[j][k,:] = deepcopy(bbot1.weights[j][k,:])
 
 class BotPool:
-    def __init__(self, bbots, fitness):
+    def __init__(self, bbots):
         self.bbots = bbots
-        # TODO Why do we keeps fitnesses seperately here and in the class?
-        self.fitness = fitness
 
     def __len__(self):
         return len(self.bbots)
@@ -25,65 +23,53 @@ class BotPool:
     def random(size, N):
         # creates a uniformly random bbotpool
         bbots = [Bot.random(size, id) for id in range(N)]
-        fitness = np.empty(N)
-        fitness.fill(np.nan)
-        return BotPool(bbots, fitness)
+        return BotPool(bbots)
 
     @staticmethod
     def zeros(N):
         # creates an empty bbotpool
         bbots = [Bot(0, id) for id in range(N)]
-        fitness = np.empty(N)
-        fitness.fill(np.nan)
-        return BotPool(bbots, fitness)
+        return BotPool(bbots)
 
     def __getitem__(self, key):
         bbots = self.bbots[key]
-        fitness = self.fitness[key]
         if hasattr(bbots, "__iter__"):  # Check if we got a list or a single object
             for n in range(len(bbots)):
                 bbots[n].id = n
-            return BotPool(bbots, fitness)
+            return BotPool(bbots)
         else:
-            return bbots, fitness
+            return bbots
 
     def __setitem__(self, key, pool):
         self.bbots[key] = pool.bbots
-        self.fitness[key] = pool.fitness
         for n in range(len(self.bbots)):
             self.bbots[n].id = n
 
     def __add__(self, pool):
         bbots = self.bbots+pool.bbots
-        fitness = np.concatenate((self.fitness,pool.fitness))
         for n in range(len(bbots)):
             bbots[n].id = n
-        return BotPool(bbots, fitness)
+        return BotPool(bbots)
 
     def compute_fitness(self, mode=None):
         # computes the fitness of the pool
         if mode == "all":
-            idx = range(len(self))
+            bbots = self.bbots
         else: # don't compute fitnesses we already know
-            nofitness = np.isnan(self.fitness)
-            idx = np.nonzero(nofitness)[0]
[…]

### 2018-12-04 project/main.py

@@ -6,8 +6,8 @@ import matplotlib.pyplot as plt
 import time
 from copy import deepcopy
 
-
-np.random.seed(7)       # because 7 is a lucky number
+# because 7 is a lucky number, our household mathemathican said so
+np.random.seed(7)       
 size = [6, 7, 2]        # size of the neuronal networks
 N = 100                 # population size
 maxgen = 100            # maximal number of generation
@@ -22,8 +22,7 @@ n_elitism = 10          # keep the best individuals
 fig = plt.figure(figsize=(10,6))
 pool = BotPool.random(size, N)
 pool.compute_fitness('all')
-fitness = [pool.fitness]
-
+fitness = [np.array([bot.fitness for bot in pool.bbots])]
 
 for gen in range(maxgen):
     gen_time = time.time()
@@ -37,7 +36,7 @@ for gen in range(maxgen):
         pool.compute_fitness()
     fitness_time = time.time() - fitness_time
     pool = pool[0:N]
-    fitness.append(pool.fitness)
+    fitness.append(np.array([bot.fitness for bot in pool.bbots]))
 
     print('Gen {}/{}; time: {:1.4f}s/{:1.4f}s'.format(
             gen, maxgen,

### 2018-12-04 project/bot.py
Make id local to the run

@@ -21,24 +21,7 @@ for f in glob.glob("blobby-1.0_fast/data/.blobby/config_*.xml"):
 class Bot:
     def __init__(self, weights, id):
         self.weights = deepcopy(weights)
-        self.id = id
         self.fitness = -np.inf
-
-    @property
-    def id(self):
-        return self._id
-
-    @id.setter
-    def id(self, id):
-        self._id = id
-        self.config_fname = 'config_{:04d}.xml'.format(id)
-        self.config_path = 'blobby-1.0_fast/data/.blobby/' + self.config_fname
-        self.nn_path = 'blobby-1.0_fast/data/.blobby/scripts/nn_{:04d}.lua'.format(id)
-        self.config_string = config_template.replace('<var name="left_script_name" value="nn"/>',
-                                                     '<var name="left_script_name" value="nn_{:04d}"/>'.format(self.id))
-        if not os.path.isfile(self.config_path):
-            with open(self.config_path, 'w') as f:
-                f.write(self.config_string)
 
     @property
     def size(self):
@@ -69,7 +52,7 @@ class Bot:
             mask = np.less(np.random.random_sample(self.weights[j].shape), rate)
             self.weights[j] += sigma*mask*np.random.randn(self.weights[j].shape[0],self.weights[j].shape[1])
 
-    def compute_fitness(self):
+    def compute_fitness(self, id):
         if test:
             # sum over squared frobenius norms of the weight matrices
             frob = 0
@@ -77,6 +60,7 @@ class Bot:
                 frob += np.sum(w**2)
             self.fitness = 1/(1+frob)
         else:
+            self.write_config(id)
             self.write_lua()
 
             score = subprocess.run(['./src/blobby', self.config_fname],
@@ -88,6 +72,16 @@ class Bot:
             score_left = int(score[1])
             score_right = int(score[2])
             self.fitness = score_left/(score_left+score_right)
+
+    def write_config(self, id):
+        self.config_fname = 'config_{:04d}.xml'.format(id)
+        self.config_path = 'blobby-1.0_fast/data/.blobby/' + self.config_fname
+        self.nn_path = 'blobby-1.0_fast/data/.blobby/scripts/nn_{:04d}.lua'.format(id)
+        self.config_string = config_template.replace('<var name="left_script_name" value="nn"/>',
+                                                     '<var name="left_script_name" value="nn_{:04d}"/>'.format(id))
+        if not os.path.isfile(self.config_path):
[…]

### 2018-12-04 project/botpool.py

@@ -61,7 +61,7 @@ class BotPool:
         chunks = np.array_split(bbots, round(len(bbots)/thread_count))
 
         for chunk in chunks:
-            threads = [threading.Thread(target=bot.compute_fitness) for bot in chunk]
+            threads = [threading.Thread(target=bot.compute_fitness, kwargs={'id':id}) for id, bot in enumerate(chunk)]
             for t in threads:
                 t.start()
             for t in threads:

### 2018-12-04 project/bot.py
Refactor botpool

@@ -19,7 +19,7 @@ for f in glob.glob("blobby-1.0_fast/data/.blobby/config_*.xml"):
     os.remove(f)
 
 class Bot:
-    def __init__(self, weights, id):
+    def __init__(self, weights):
         self.weights = deepcopy(weights)
         self.fitness = -np.inf
 
@@ -38,12 +38,12 @@ class Bot:
         return len(self.size)
 
     @staticmethod
-    def random(size, id):
+    def random(size):
         # creates a uniformly randon bbot
         weights = []
         for j in range(len(size)-1):
             weights.append(np.random.rand(size[j+1], size[j]+1)-0.5)
-        return Bot(weights, id)
+        return Bot(weights)
 
     def mutate_gaussian(self, rate, sigma):
         self.fitness = -np.inf

### 2018-12-04 project/botpool.py

@@ -1,110 +1,0 @@
-from bot import Bot
-import numpy as np
-import threading
-from copy import deepcopy
-
-def crossover_mix_nodes(bbot1, bbot2, offspring1, offspring2, rate):
-    offspring1.weights = deepcopy(bbot1.weights)
-    offspring2.weights = deepcopy(bbot2.weights)
-    for j in range(bbot1.nlayers-1):
-        for k in range(bbot1.size[j+1]):
-            if np.random.rand() < rate:
-                offspring1.weights[j][k,:] = deepcopy(bbot2.weights[j][k,:])
-                offspring2.weights[j][k,:] = deepcopy(bbot1.weights[j][k,:])
-
-class BotPool:
-    def __init__(self, bbots):
-        self.bbots = bbots
-
-    def __len__(self):
-        return len(self.bbots)
-
-    @staticmethod
-    def random(size, N):
-        # creates a uniformly random bbotpool
-        bbots = [Bot.random(size, id) for id in range(N)]
-        return BotPool(bbots)
-
-    @staticmethod
-    def zeros(N):
-        # creates an empty bbotpool
-        bbots = [Bot(0, id) for id in range(N)]
-        return BotPool(bbots)
-
-    def __getitem__(self, key):
-        bbots = self.bbots[key]
-        if hasattr(bbots, "__iter__"):  # Check if we got a list or a single object
-            for n in range(len(bbots)):
-                bbots[n].id = n
-            return BotPool(bbots)
-        else:
-            return bbots
-
-    def __setitem__(self, key, pool):
-        self.bbots[key] = pool.bbots
-        for n in range(len(self.bbots)):
-            self.bbots[n].id = n
-
-    def __add__(self, pool):
-        bbots = self.bbots+pool.bbots
-        for n in range(len(bbots)):
-            bbots[n].id = n
-        return BotPool(bbots)
-
-    def compute_fitness(self, mode=None):
-        # computes the fitness of the pool
-        if mode == "all":
-            bbots = self.bbots
-        else: # don't compute fitnesses we already know
-            bbots = [bot for bot in self.bbots if bot.fitness == -np.inf]
-        thread_count = 50 # maximal parallel threads
-        chunks = np.array_split(bbots, round(len(bbots)/thread_count))
-
-        for chunk in chunks:
-            threads = [threading.Thread(target=bot.compute_fitness, kwargs={'id':id}) for id, bot in enumerate(chunk)]
[…]

### 2018-12-04 project/genetic_operators.py

@@ -1,0 +1,44 @@
+from copy import deepcopy
+from bot import Bot
+import numpy as np
+
+def crossover_mix_nodes(bbot1, bbot2, offspring1, offspring2, rate):
+    offspring1.weights = deepcopy(bbot1.weights)
+    offspring2.weights = deepcopy(bbot2.weights)
+    for j in range(bbot1.nlayers-1):
+        for k in range(bbot1.size[j+1]):
+            if np.random.rand() < rate:
+                offspring1.weights[j][k,:] = deepcopy(bbot2.weights[j][k,:])
+                offspring2.weights[j][k,:] = deepcopy(bbot1.weights[j][k,:])
+
+def crossover_roulette_wheel(bots, N, p_crossover, rate):
+    N = round(N/2)*2
+    fitnesses = np.array([bot.fitness for bot in bots])
+    if np.sum(fitnesses) == 0:
+        dist = np.arange(len(bots))/len(bots) # uniform wheel
+    else:
+        fbar = fitnesses + 1
+        dist = np.cumsum(fbar/np.sum(fbar))
+
+    idx = []
+    for x in np.random.rand(N):
+        for n in range(len(bots)):
+            if x < dist[n]:
+                idx.append(n)
+                break
+
+    offsprings = []
+
+    for n in range(int(N/2)):
+        if np.random.rand() < p_crossover:
+            offsprings.extend([Bot(None), Bot(None)])
+            crossover_mix_nodes(bots[idx[2*n]], bots[idx[2*n+1]], offsprings[-2], offsprings[-1], rate)
+        else:
+            offsprings.extend([bots[idx[2*n]], bots[idx[2*n+1]]])
+
+    return offsprings
+
+def mutate_gaussian(bots, p_mutation, rate, sigma):
+    for n in range(len(bots)):
+        if np.random.rand() < p_mutation:
+            bots[n].mutate_gaussian(rate, sigma)

### 2018-12-04 project/main.py

@@ -1,47 +1,78 @@
 #!/bin/bash
-from botpool import BotPool
 from little_functions import plot
 import numpy as np
 import matplotlib.pyplot as plt
 import time
 from copy import deepcopy
+import threading
+from genetic_operators import crossover_roulette_wheel, mutate_gaussian
+from bot import Bot
+import math
 
-# because 7 is a lucky number, our household mathemathican said so
-np.random.seed(7)       
-size = [6, 7, 2]        # size of the neuronal networks
-N = 100                 # population size
-maxgen = 100            # maximal number of generation
-N_offspring = 150
-p_crossover = 0.9       # crossove parameters
-crossover_rate = 0.5
-p_mutation = 0.5
-mutation_rate = 0.5     # mutation parameters
-mutation_sigma = 0.01
-n_elitism = 10          # keep the best individuals
+def inverse_fitness_sort(pool):
+    pool.sort(key=lambda bot: -bot.fitness)
 
-fig = plt.figure(figsize=(10,6))
-pool = BotPool.random(size, N)
-pool.compute_fitness('all')
-fitness = [np.array([bot.fitness for bot in pool.bbots])]
+def compute_fitness(pool, mode=None):
+    # computes the fitness of the pool
+    if mode == "all":
+        subpool = pool
+    else: # don't compute fitnesses we already know
+        subpool = [bot for bot in pool if bot.fitness == -np.inf]
+    thread_count = 50 # maximal parallel threads
+    chunks = np.array_split(subpool, math.ceil(len(subpool)/thread_count))
 
-for gen in range(maxgen):
-    gen_time = time.time()
-    offspring = pool.crossover_roulette_wheel(N_offspring, p_crossover, crossover_rate)
-    offspring.mutate_gaussian(p_mutation, mutation_rate, mutation_sigma)
-    pool = deepcopy(pool[0:n_elitism]+offspring)
-    fitness_time = time.time()
-    if gen%5 == 0:
-        pool.compute_fitness('all')
-    else:
-        pool.compute_fitness()
-    fitness_time = time.time() - fitness_time
-    pool = pool[0:N]
-    fitness.append(np.array([bot.fitness for bot in pool.bbots]))
+    for chunk in chunks:
+        threads = [threading.Thread(target=bot.compute_fitness, kwargs={'id':id}) for id, bot in enumerate(chunk)]
+        for t in threads:
+            t.start()
+        for t in threads:
+            t.join()
[…]

### 2018-12-04 project/main.py
Cosmetic changes

@@ -27,11 +27,6 @@ def compute_fitness(pool, mode=None):
             t.start()
         for t in threads:
             t.join()
-    inverse_fitness_sort(pool)
-
-def write_best(pool):
-    inverse_fitness_sort(pool)
-    pool[0].write_lua('nn_max.lua')
 
 def main():
     # because 7 is a lucky number, our household mathemathican said so
@@ -63,6 +58,8 @@ def main():
         else:
             compute_fitness(pool)
         fitness_time = time.time() - fitness_time
+        inverse_fitness_sort(pool)
+        pool[0].write_lua('nn_max.lua')
         pool = pool[0:N]
         fitness.append(np.array([bot.fitness for bot in pool]))
 
@@ -70,7 +67,6 @@ def main():
                 gen, maxgen,
                 fitness_time, time.time() - gen_time))
         plot(fig, fitness)
-        write_best(pool)
 
     plt.show()
 

### 2018-12-05 project/main.py
Add a missing sort after the 0th gen

@@ -45,6 +45,7 @@ def main():
     fig = plt.figure(figsize=(10,6))
     pool = [Bot.random(size) for _ in range(N)]
     compute_fitness(pool, 'all')
+    inverse_fitness_sort(pool)
     fitness = [np.array([bot.fitness for bot in pool])]
 
     for gen in range(maxgen):

### 2018-12-05 project/genetic_operators.py
Correct copying of parents for crossover

@@ -34,7 +34,7 @@ def crossover_roulette_wheel(bots, N, p_crossover, rate):
             offsprings.extend([Bot(None), Bot(None)])
             crossover_mix_nodes(bots[idx[2*n]], bots[idx[2*n+1]], offsprings[-2], offsprings[-1], rate)
         else:
-            offsprings.extend([bots[idx[2*n]], bots[idx[2*n+1]]])
+            offsprings.extend([deepcopy(bots[idx[2*n]]), deepcopy(bots[idx[2*n+1]])])
 
     return offsprings
 

### 2018-12-05 project/little_functions.py
outsourced the plotting from main.py to plot_fitness.py

@@ -1,39 +1,0 @@
-import matplotlib.pyplot as plt
-import numpy as np
-
-
-def plot(fig, fitness):
-    gen = len(fitness)
-    if gen == 1:
-        return
-    fig.clf()
-
-    ax = plt.axes()
-    ax.plot(np.arange(gen), np.mean(fitness, axis = 1), 'k')
-    ax.plot(np.arange(gen), np.max(fitness, axis = 1), 'k')
-    ax.plot(np.arange(gen), np.min(fitness, axis = 1), 'k')
-
-    yres = 20
-    y_grid = np.linspace(0, 1, yres)
-    zi = [ [ np.sum(np.less_equal(y_grid[j],fitness[k])*np.less(fitness[k],y_grid[j+1])) for j in range(len(y_grid)-1)] for k in range(gen) ]
-
-    zi = np.transpose(np.array(zi))
-    zi = np.concatenate((zi, np.zeros([1,gen])))
-    xi, yi = np.meshgrid(np.arange(0,gen)-0.5, y_grid)
-    tmp = ax.pcolor(xi, yi, zi, cmap=plt.cm.Oranges)
-    cb = plt.colorbar(tmp)
-    cb.set_label('% of population')
-    tmp.set_clim(vmin=0, vmax=20)
-
-    ax.set_xlabel('generation')
-    ax.set_ylabel('fitness')
-    ax.set_xlim([0,gen-1.5])
-    ax.set_ylim([0,1])
-
-    font = {'family' : 'normal',
-            'size'   : 22}
-    plt.rc('font', **font)
-    plt.tight_layout()
-
-    plt.show(block=False)
-    plt.pause(0.1)

### 2018-12-05 project/main.py

@@ -1,5 +1,4 @@
 #!/bin/bash
-from little_functions import plot
 import numpy as np
 import matplotlib.pyplot as plt
 import time
@@ -8,6 +7,7 @@ import threading
 from genetic_operators import crossover_roulette_wheel, mutate_gaussian
 from bot import Bot
 import math
+import pickle
 
 def inverse_fitness_sort(pool):
     pool.sort(key=lambda bot: -bot.fitness)
@@ -28,6 +28,10 @@ def compute_fitness(pool, mode=None):
         for t in threads:
             t.join()
 
+def save_fitness(fitness):
+    with open('fitness.pkl', 'wb') as f:
+        pickle.dump(fitness, f)
+
 def main():
     # because 7 is a lucky number, our household mathemathican said so
     np.random.seed(7)
@@ -42,13 +46,12 @@ def main():
     mutation_sigma = 0.01
     n_elitism = 10          # keep the best individuals
 
-    fig = plt.figure(figsize=(10,6))
     pool = [Bot.random(size) for _ in range(N)]
     compute_fitness(pool, 'all')
     inverse_fitness_sort(pool)
     fitness = [np.array([bot.fitness for bot in pool])]
 
-    for gen in range(maxgen):
+    for gen in range(maxgen+1):
         gen_time = time.time()
         offspring = crossover_roulette_wheel(pool, n_offspring, p_crossover, crossover_rate)
         mutate_gaussian(offspring, p_mutation, mutation_rate, mutation_sigma)
@@ -64,12 +67,14 @@ def main():
         pool = pool[0:N]
         fitness.append(np.array([bot.fitness for bot in pool]))
 
-        print('Gen {}/{}; time: {:1.4f}s/{:1.4f}s'.format(
+        print('Gen {}/{}; time: {:1.4f}s/{:1.4f}s; min: {:1.4f}; mean: {:1.4f}; max: {:1.4f}'.format(
                 gen, maxgen,
-                fitness_time, time.time() - gen_time))
-        plot(fig, fitness)
+                fitness_time, time.time() - gen_time,
+                np.min(fitness[gen]),
+                np.mean(fitness[gen]),
+                np.max(fitness[gen])))
 
-    plt.show()
+    save_fitness(fitness)
 
 if __name__ == '__main__':
     main()

### 2018-12-05 project/plot_fitness.py

@@ -1,0 +1,39 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import pickle
+
+with open('fitness.pkl', 'rb') as f:
+    fitness = pickle.load(f)
+
+gen = len(fitness)
+
+fig = plt.figure(figsize=(10,6))
+
+ax = plt.axes()
+ax.plot(np.arange(gen), np.mean(fitness, axis = 1), 'k')
+ax.plot(np.arange(gen), np.max(fitness, axis = 1), 'k')
+ax.plot(np.arange(gen), np.min(fitness, axis = 1), 'k')
+
+yres = 20
+y_grid = np.linspace(0, 1, yres)
+zi = [ [ np.sum(np.less_equal(y_grid[j],fitness[k])*np.less(fitness[k],y_grid[j+1])) for j in range(len(y_grid)-1)] for k in range(gen) ]
+
+zi = np.transpose(np.array(zi))
+zi = np.concatenate((zi, np.zeros([1,gen])))
+xi, yi = np.meshgrid(np.arange(0,gen)-0.5, y_grid)
+tmp = ax.pcolor(xi, yi, zi, cmap=plt.cm.Oranges)
+cb = plt.colorbar(tmp)
+cb.set_label('% of population')
+tmp.set_clim(vmin=0, vmax=20)
+
+ax.set_xlabel('generation')
+ax.set_ylabel('fitness')
+ax.set_xlim([0,gen-1.5])
+ax.set_ylim([0,1])
+
+font = {'family' : 'normal',
+        'size'   : 22}
+plt.rc('font', **font)
+plt.tight_layout()
+
+plt.show()

### 2018-12-05 project/main.py
cosmetic change

@@ -67,7 +67,7 @@ def main():
         pool = pool[0:N]
         fitness.append(np.array([bot.fitness for bot in pool]))
 
-        print('Gen {}/{}; time: {:1.4f}s/{:1.4f}s; min: {:1.4f}; mean: {:1.4f}; max: {:1.4f}'.format(
+        print('Gen {}/{}; \ttime: {:1.4f}s/{:1.4f}s; \tmin = {:1.4f}; \tmean = {:1.4f}; \tmax = {:1.4f}'.format(
                 gen, maxgen,
                 fitness_time, time.time() - gen_time,
                 np.min(fitness[gen]),

### 2018-12-05 project/bot.py
cosmetics while searching for a bug - have not found any

@@ -25,7 +25,7 @@ class Bot:
 
     @property
     def size(self):
-        # returns the size of the neuronal network
+        # returns the size of the neural network
         size = []
         for w in self.weights:
             size.append(np.shape(w)[1]-1)
@@ -39,7 +39,7 @@ class Bot:
 
     @staticmethod
     def random(size):
-        # creates a uniformly randon bbot
+        # creates a uniformly random bbot
         weights = []
         for j in range(len(size)-1):
             weights.append(np.random.rand(size[j+1], size[j]+1)-0.5)
@@ -50,7 +50,7 @@ class Bot:
         # adds Gaussian noise with deviation sigma to rate weights (0.5 for half of them)
         for j in range(self.nlayers-1):
             mask = np.less(np.random.random_sample(self.weights[j].shape), rate)
-            self.weights[j] += sigma*mask*np.random.randn(self.weights[j].shape[0],self.weights[j].shape[1])
+            self.weights[j] += sigma*mask*np.random.randn(self.size[j+1],self.size[j]+1)
 
     def compute_fitness(self, id):
         if test:

### 2018-12-05 project/bot.py
added option to self adapt the mutation parameters

@@ -19,9 +19,12 @@ for f in glob.glob("blobby-1.0_fast/data/.blobby/config_*.xml"):
     os.remove(f)
 
 class Bot:
-    def __init__(self, weights):
-        self.weights = deepcopy(weights)
-        self.fitness = -np.inf
+    def __init__(self, weights, mutation_p, mutation_rate, mutation_sigma):
+        self.weights        = deepcopy(weights)
+        self.fitness        = -np.inf
+        self.mutation_p     = mutation_p
+        self.mutation_rate  = mutation_rate
+        self.mutation_sigma = mutation_sigma
 
     @property
     def size(self):
@@ -43,11 +46,24 @@ class Bot:
         weights = []
         for j in range(len(size)-1):
             weights.append(np.random.rand(size[j+1], size[j]+1)-0.5)
-        return Bot(weights)
+        mutation_p = np.random.rand()
+        mutation_rate = np.random.rand()
+        mutation_sigma = 0.05*np.random.rand()
+        return Bot(weights, mutation_p, mutation_rate, mutation_sigma)
 
-    def mutate_gaussian(self, rate, sigma):
+    def mutate_gaussian(self, *args):
         self.fitness = -np.inf
         # adds Gaussian noise with deviation sigma to rate weights (0.5 for half of them)
+        if args[0] == 'self_adapt':
+            self.mutation_p += 0.1*np.random.randn()
+            self.mutation_rate += 0.1*np.random.randn()
+            self.mutation_sigma += 0.01*np.random.randn()
+            rate = self.mutation_rate
+            sigma = self.mutation_rate
+        else:
+            rate = args[0]
+            sigma = args[1]
+
         for j in range(self.nlayers-1):
             mask = np.less(np.random.random_sample(self.weights[j].shape), rate)
             self.weights[j] += sigma*mask*np.random.randn(self.size[j+1],self.size[j]+1)

### 2018-12-05 project/genetic_operators.py

@@ -3,8 +3,14 @@ from bot import Bot
 import numpy as np
 
 def crossover_mix_nodes(bbot1, bbot2, rate):
-    offspring1 = Bot(bbot1.weights)
-    offspring2 = Bot(bbot2.weights)
+    offspring1 = deepcopy(bbot1)
+    offspring2 = deepcopy(bbot2)
+    offspring1.mutation_p = (1-rate)*bbot1.mutation_p+rate*bbot2.mutation_p
+    offspring1.mutation_rate = (1-rate)*bbot1.mutation_p+rate*bbot2.mutation_p
+    offspring1.mutation_sigma = (1-rate)*bbot1.mutation_p+rate*bbot2.mutation_p
+    offspring2.mutation_p = (1-rate)*bbot2.mutation_p+rate*bbot1.mutation_p
+    offspring2.mutation_rate = (1-rate)*bbot2.mutation_p+rate*bbot1.mutation_p
+    offspring2.mutation_sigma = (1-rate)*bbot2.mutation_p+rate*bbot1.mutation_p
     for j in range(bbot1.nlayers-1):
         for k in range(bbot1.size[j+1]):
             if np.random.rand() < rate:
@@ -13,7 +19,7 @@ def crossover_mix_nodes(bbot1, bbot2, rate):
 
     return offspring1, offspring2
 
-def crossover_roulette_wheel(bots, N, p_crossover, rate):
+def crossover_roulette_wheel(bots, N, crossover_p, rate):
     N = round(N/2)*2
     fitnesses = np.array([bot.fitness for bot in bots])
     if np.sum(fitnesses) == 0:
@@ -31,14 +37,22 @@ def crossover_roulette_wheel(bots, N, p_crossover, rate):
 
     offsprings = []
     for n in range(int(N/2)):
-        if np.random.rand() < p_crossover:
+        if np.random.rand() < crossover_p:
             offsprings.extend(crossover_mix_nodes(bots[idx[2*n]], bots[idx[2*n+1]], rate))
         else:
             offsprings.extend([deepcopy(bots[idx[2*n]]), deepcopy(bots[idx[2*n+1]])])
 
     return offsprings
 
-def mutate_gaussian(bots, p_mutation, rate, sigma):
-    for n in range(len(bots)):
-        if np.random.rand() < p_mutation:
-            bots[n].mutate_gaussian(rate, sigma)
+def mutate_gaussian(bots, *args):
+    if args[0] == 'self_adapt':
+        for n in range(len(bots)):
[…]

### 2018-12-05 project/main.py

@@ -39,9 +39,9 @@ def main():
     N = 100                 # population size
     maxgen = 100            # maximal number of generation
     n_offspring = 150
-    p_crossover = 0.9       # crossove parameters
+    crossover_p = 0.9       # crossove parameters
     crossover_rate = 0.5
-    p_mutation = 0.5
+    mutation_p = 0.5
     mutation_rate = 0.5     # mutation parameters
     mutation_sigma = 0.01
     n_elitism = 10          # keep the best individuals
@@ -53,8 +53,9 @@ def main():
 
     for gen in range(maxgen+1):
         gen_time = time.time()
-        offspring = crossover_roulette_wheel(pool, n_offspring, p_crossover, crossover_rate)
-        mutate_gaussian(offspring, p_mutation, mutation_rate, mutation_sigma)
+        offspring = crossover_roulette_wheel(pool, n_offspring, crossover_p, crossover_rate)
+#        mutate_gaussian(offspring, mutation_p, mutation_rate, mutation_sigma)
+        mutate_gaussian(offspring, 'self_adapt')
         pool = pool[0:n_elitism] + offspring
         fitness_time = time.time()
         if gen%5 == 0:

### 2018-12-05 project/bot.py
bug fix and little parameter tuning

@@ -48,7 +48,7 @@ class Bot:
             weights.append(np.random.rand(size[j+1], size[j]+1)-0.5)
         mutation_p = np.random.rand()
         mutation_rate = np.random.rand()
-        mutation_sigma = 0.05*np.random.rand()
+        mutation_sigma = 0.02*np.random.rand()
         return Bot(weights, mutation_p, mutation_rate, mutation_sigma)
 
     def mutate_gaussian(self, *args):

### 2018-12-05 project/genetic_operators.py

@@ -6,11 +6,11 @@ def crossover_mix_nodes(bbot1, bbot2, rate):
     offspring1 = deepcopy(bbot1)
     offspring2 = deepcopy(bbot2)
     offspring1.mutation_p = (1-rate)*bbot1.mutation_p+rate*bbot2.mutation_p
-    offspring1.mutation_rate = (1-rate)*bbot1.mutation_p+rate*bbot2.mutation_p
-    offspring1.mutation_sigma = (1-rate)*bbot1.mutation_p+rate*bbot2.mutation_p
     offspring2.mutation_p = (1-rate)*bbot2.mutation_p+rate*bbot1.mutation_p
-    offspring2.mutation_rate = (1-rate)*bbot2.mutation_p+rate*bbot1.mutation_p
-    offspring2.mutation_sigma = (1-rate)*bbot2.mutation_p+rate*bbot1.mutation_p
+    offspring1.mutation_rate = (1-rate)*bbot1.mutation_rate+rate*bbot2.mutation_rate
+    offspring2.mutation_rate = (1-rate)*bbot2.mutation_rate+rate*bbot1.mutation_rate
+    offspring1.mutation_sigma = (1-rate)*bbot1.mutation_sigma+rate*bbot2.mutation_sigma
+    offspring2.mutation_sigma = (1-rate)*bbot2.mutation_sigma+rate*bbot1.mutation_sigma
     for j in range(bbot1.nlayers-1):
         for k in range(bbot1.size[j+1]):
             if np.random.rand() < rate:

### 2018-12-05 project/bot.py
added two self-adapting methods and did some parameter tuning

@@ -19,12 +19,9 @@ for f in glob.glob("blobby-1.0_fast/data/.blobby/config_*.xml"):
     os.remove(f)
 
 class Bot:
-    def __init__(self, weights, mutation_p, mutation_rate, mutation_sigma):
+    def __init__(self, weights):
         self.weights        = deepcopy(weights)
         self.fitness        = -np.inf
-        self.mutation_p     = mutation_p
-        self.mutation_rate  = mutation_rate
-        self.mutation_sigma = mutation_sigma
 
     @property
     def size(self):
@@ -41,32 +38,42 @@ class Bot:
         return len(self.size)
 
     @staticmethod
-    def random(size):
+    def random(size, mode = None):
         # creates a uniformly random bbot
         weights = []
         for j in range(len(size)-1):
             weights.append(np.random.rand(size[j+1], size[j]+1)-0.5)
-        mutation_p = np.random.rand()
-        mutation_rate = np.random.rand()
-        mutation_sigma = 0.02*np.random.rand()
-        return Bot(weights, mutation_p, mutation_rate, mutation_sigma)
+        bbot = Bot(weights)
+        bbot.mode = mode
+        if mode == 'self_adapt1':
+            bbot.mutation_rate = np.random.rand()/2
+            bbot.mutation_sigma = 0.05*np.random.rand()
+        if mode == 'self_adapt2':
+            bbot.mutation_sigma = []
+            for j in range(len(size)-1):
+                bbot.mutation_sigma.append(0.05*np.random.randn(size[j+1], size[j]+1))
+
+        return bbot
 
     def mutate_gaussian(self, *args):
         self.fitness = -np.inf
         # adds Gaussian noise with deviation sigma to rate weights (0.5 for half of them)
-        if args[0] == 'self_adapt':
-            self.mutation_p += 0.1*np.random.randn()
-            self.mutation_rate += 0.1*np.random.randn()
-            self.mutation_sigma += 0.01*np.random.randn()
-            rate = self.mutation_rate
-            sigma = self.mutation_rate
+        if self.mode == 'self_adapt1':
+            self.mutation_sigma *= np.exp(0.05*np.random.randn())
+            self.mutation_rate *= np.exp(0.1*np.random.randn())
+            for j in range(self.nlayers-1):
+                mask = np.less(np.random.random_sample(self.weights[j].shape), self.mutation_rate)
[…]

### 2018-12-05 project/ea_adaptive1.py

@@ -1,0 +1,80 @@
+#!/bin/bash
+import numpy as np
+import matplotlib.pyplot as plt
+import time
+from copy import deepcopy
+import threading
+from genetic_operators import crossover_roulette_wheel, mutate_gaussian
+from bot import Bot
+import math
+import pickle
+
+def inverse_fitness_sort(pool):
+    pool.sort(key=lambda bot: -bot.fitness)
+
+def compute_fitness(pool, mode=None):
+    # computes the fitness of the pool
+    if mode == "all":
+        subpool = pool
+    else: # don't compute fitnesses we already know
+        subpool = [bot for bot in pool if bot.fitness == -np.inf]
+    thread_count = 100 # maximal parallel threads
+    chunks = np.array_split(subpool, math.ceil(len(subpool)/thread_count))
+
+    for chunk in chunks:
+        threads = [threading.Thread(target=bot.compute_fitness, kwargs={'id':id}) for id, bot in enumerate(chunk)]
+        for t in threads:
+            t.start()
+        for t in threads:
+            t.join()
+
+def save_fitness(fitness):
+    with open('fitness.pkl', 'wb') as f:
+        pickle.dump(fitness, f)
+
+def main():
+    # because 7 is a lucky number, our household mathemathican said so
+    np.random.seed(7)
+    size = [6, 7, 2]        # size of the neuronal networks
+    N = 100                 # population size
+    maxgen = 200            # maximal number of generation
+    n_offspring = 150
+    crossover_p = 0.9       # crossove parameters
+    crossover_rate = 0.3
+    fbar = lambda x: (x+0.5)**3
+    mutation_p = 0.9
+    n_elitism = 2           # keep the best individuals
+
+    pool = [Bot.random(size, 'self_adapt1') for _ in range(N)]
+    compute_fitness(pool, 'all')
+    inverse_fitness_sort(pool)
+    fitness = [np.array([bot.fitness for bot in pool])]
+
+    for gen in range(maxgen+1):
+        gen_time = time.time()
+        offspring = crossover_roulette_wheel(pool, n_offspring, crossover_p, crossover_rate, fbar)
+        mutate_gaussian(offspring, mutation_p)
+        pool = pool[0:n_elitism] + offspring
+        fitness_time = time.time()
+        if gen%5 == 0:
+            compute_fitness(pool, 'all')
+        else:
+            compute_fitness(pool)
[…]

### 2018-12-05 project/ea_adaptive2.py

@@ -18,7 +18,7 @@ def compute_fitness(pool, mode=None):
         subpool = pool
     else: # don't compute fitnesses we already know
         subpool = [bot for bot in pool if bot.fitness == -np.inf]
-    thread_count = 50 # maximal parallel threads
+    thread_count = 100 # maximal parallel threads
     chunks = np.array_split(subpool, math.ceil(len(subpool)/thread_count))
 
     for chunk in chunks:
@@ -37,25 +37,23 @@ def main():
     np.random.seed(7)
     size = [6, 7, 2]        # size of the neuronal networks
     N = 100                 # population size
-    maxgen = 100            # maximal number of generation
+    maxgen = 200            # maximal number of generation
     n_offspring = 150
     crossover_p = 0.9       # crossove parameters
-    crossover_rate = 0.5
-    mutation_p = 0.5
-    mutation_rate = 0.5     # mutation parameters
-    mutation_sigma = 0.01
-    n_elitism = 10          # keep the best individuals
+    crossover_rate = 0.4
+    fbar = lambda x: (x+0.3)**3
+    mutation_p = 0.9
+    n_elitism = 2           # keep the best individuals
 
-    pool = [Bot.random(size) for _ in range(N)]
+    pool = [Bot.random(size, 'self_adapt2') for _ in range(N)]
     compute_fitness(pool, 'all')
     inverse_fitness_sort(pool)
     fitness = [np.array([bot.fitness for bot in pool])]
 
     for gen in range(maxgen+1):
         gen_time = time.time()
-        offspring = crossover_roulette_wheel(pool, n_offspring, crossover_p, crossover_rate)
-#        mutate_gaussian(offspring, mutation_p, mutation_rate, mutation_sigma)
-        mutate_gaussian(offspring, 'self_adapt')
+        offspring = crossover_roulette_wheel(pool, n_offspring, crossover_p, crossover_rate, fbar)
+        mutate_gaussian(offspring, mutation_p)
         pool = pool[0:n_elitism] + offspring
         fitness_time = time.time()
         if gen%5 == 0:
@@ -74,8 +72,8 @@ def main():
                 np.min(fitness[gen]),
                 np.mean(fitness[gen]),
                 np.max(fitness[gen])))
+        save_fitness(fitness)
 
-    save_fitness(fitness)
 
 if __name__ == '__main__':
     main()

### 2018-12-05 project/ea_standart.py

@@ -1,0 +1,80 @@
+#!/bin/bash
+import numpy as np
+import matplotlib.pyplot as plt
+import time
+from copy import deepcopy
+import threading
+from genetic_operators import crossover_roulette_wheel, mutate_gaussian
+from bot import Bot
+import math
+import pickle
+
+def inverse_fitness_sort(pool):
+    pool.sort(key=lambda bot: -bot.fitness)
+
+def compute_fitness(pool, mode=None):
+    # computes the fitness of the pool
+    if mode == "all":
+        subpool = pool
+    else: # don't compute fitnesses we already know
+        subpool = [bot for bot in pool if bot.fitness == -np.inf]
+    thread_count = 100 # maximal parallel threads
+    chunks = np.array_split(subpool, math.ceil(len(subpool)/thread_count))
+
+    for chunk in chunks:
+        threads = [threading.Thread(target=bot.compute_fitness, kwargs={'id':id}) for id, bot in enumerate(chunk)]
+        for t in threads:
+            t.start()
+        for t in threads:
+            t.join()
+
+def save_fitness(fitness):
+    with open('fitness.pkl', 'wb') as f:
+        pickle.dump(fitness, f)
+
+def main():
+    # because 7 is a lucky number, our household mathemathican said so
+    np.random.seed(7)
+    size = [6, 7, 2]        # size of the neuronal networks
+    N = 100                 # population size
+    maxgen = 100            # maximal number of generation
+    n_offspring = 150
+    crossover_p = 0.9       # crossove parameters
+    crossover_rate = 0.5
+    fbar = lambda x: (x+1)**2
+    mutation_p = 0.5        # mutation parameters
+    mutation_rate = 0.5
+    mutation_sigma = 0.01
+    n_elitism = 10          # keep the best individuals
+
+    pool = [Bot.random(size) for _ in range(N)]
+    compute_fitness(pool, 'all')
+    inverse_fitness_sort(pool)
+    fitness = [np.array([bot.fitness for bot in pool])]
+
+    for gen in range(maxgen+1):
+        gen_time = time.time()
+        offspring = crossover_roulette_wheel(pool, n_offspring, crossover_p, crossover_rate, fbar)
+        mutate_gaussian(offspring, mutation_p, mutation_rate, mutation_sigma)
+        pool = pool[0:n_elitism] + offspring
+        fitness_time = time.time()
+        if gen%5 == 0:
[…]

### 2018-12-05 project/genetic_operators.py

@@ -5,12 +5,17 @@ import numpy as np
 def crossover_mix_nodes(bbot1, bbot2, rate):
     offspring1 = deepcopy(bbot1)
     offspring2 = deepcopy(bbot2)
-    offspring1.mutation_p = (1-rate)*bbot1.mutation_p+rate*bbot2.mutation_p
-    offspring2.mutation_p = (1-rate)*bbot2.mutation_p+rate*bbot1.mutation_p
-    offspring1.mutation_rate = (1-rate)*bbot1.mutation_rate+rate*bbot2.mutation_rate
-    offspring2.mutation_rate = (1-rate)*bbot2.mutation_rate+rate*bbot1.mutation_rate
-    offspring1.mutation_sigma = (1-rate)*bbot1.mutation_sigma+rate*bbot2.mutation_sigma
-    offspring2.mutation_sigma = (1-rate)*bbot2.mutation_sigma+rate*bbot1.mutation_sigma
+    offspring1.fitness = -np.inf
+    offspring2.fitness = -np.inf
+    if bbot1.mode == 'self_adapt1':
+        [offspring1.mutation_rate, offspring1.mutation_sigma] = (1-rate)*np.array([bbot1.mutation_rate, bbot1.mutation_sigma])+rate*np.array([bbot2.mutation_rate, bbot2.mutation_sigma])
+        [offspring2.mutation_rate, offspring2.mutation_sigma] = (1-rate)*np.array([bbot2.mutation_rate, bbot2.mutation_sigma])+rate*np.array([bbot1.mutation_rate, bbot1.mutation_sigma])
+    if bbot1.mode == 'self_adapt2':
+        offspring1.mutation_sigma = []
+        offspring2.mutation_sigma = []
+        for s1, s2 in zip(bbot1.mutation_sigma, bbot2.mutation_sigma):
+            offspring1.mutation_sigma.append((1-rate)*s1+rate*s2)
+            offspring2.mutation_sigma.append((1-rate)*s2+rate*s1)
     for j in range(bbot1.nlayers-1):
         for k in range(bbot1.size[j+1]):
             if np.random.rand() < rate:
@@ -19,14 +24,13 @@ def crossover_mix_nodes(bbot1, bbot2, rate):
 
     return offspring1, offspring2
 
-def crossover_roulette_wheel(bots, N, crossover_p, rate):
+def crossover_roulette_wheel(bots, N, crossover_p, rate, fbar):
     N = round(N/2)*2
     fitnesses = np.array([bot.fitness for bot in bots])
[…]

### 2018-12-16 project/main.py
tried different bots

@@ -54,16 +54,29 @@ def main():
     crossover_rate = 0.5
     mutation_p = 0.7        # mutation parameters
 
-#    trainers = ['trainer_5t']; gens = [200]
-#    trainers = ['trainer'];    gens = [200]
-#    trainers = ['reduced_5t']; gens = [200]
 
-    trainers = ['trainer_5t'] # warmup
-    gens = [20]
-    trainers.extend(['reduced_5t', 'reduced', 'trainer', 'reduced']*4)
-    gens.extend([15, 30, 15, 30]*4)
-    trainers.append('reduced')
-    gens.append(50)
+#    trainers = ['reduced'];
+#    trainers = ['com_11'];
+#    trainers = ['gintonicV9'];
+#    trainers = ['hyp014'];
+#    trainers = ['Union'];
+#    trainers = ['trainer'];
+
+#    trainers = ['reduced_5t'];
+#    trainers = ['com_11_5t'];
+#    trainers = ['gintonicV9_5t'];
+#    trainers = ['hyp014_5t'];
+#    trainers = ['Union_5t'];
+    trainers = ['trainer_5t'];
+
+    gens = [200]
+
+
+
+#    trainers = ['trainer_5t'] # warmup
+#    gens = [20]
+#    trainers.extend(['com_11_5t', 'gintonicV9_5t', 'hyp014_5t', 'Union_5t', 'reduced_5t'])
+#    gens.extend([15]*5)
 
     if args.self_adapt_1:
         print('Using self adapt type 1')
@@ -123,7 +136,8 @@ def main():
                 np.min(fitness[-1]),
                 np.mean(fitness[-1]),
                 np.max(fitness[-1])))
-            save_fitness(fitness)
+            if ~gen%10:
+                save_fitness(fitness)
             gen += 1
 
 

### 2018-12-16 documentation/paper.tex
started documentation

--- 2018-12-04 project_proposal/pp.tex
@@
 %}}}
-
-% a todo command===============================================================
+%{{{ a todo command
 \newcounter{todocounter}
@@
  \marginpar{\fcolorbox{black}{yellow}{\footnotesize\textbf{todo}}
- \ifthenelse{\equal{#1}{noisnotdefined}}{}{\textcolor{black}{\newline\tiny #1}}}
+ \ifthenelse{\equal{#1}{noisnotdefined}}{}{\textcolor{black}{\newline\tiny #1}} }
  \textbf{\ifthenelse{\equal{#2}{.}}
-   {\fcolorbox{blue}{white}{\textcolor{blue}{$\maltese$}}}{{\textcolor{blue}{#2}}}}
- \refstepcounter{todocounter}}
-
-%===============================================================================
+   {\fcolorbox{blue}{white}{\textcolor{blue}{$\maltese$} } }{ {\textcolor{blue}{#2} } } }
+ \refstepcounter{todocounter} }
+%}}}
 
 \date{}
-\title{Project Proposal for Genetic Algorithms and Evolutionary Programming}
+\title{todo: think about good title}
 \author{Felix Bartel and D\'avid Kerekes}
@@
 
-\section{Problem Overview}
+\section{Introduction}
 
@@
 
-\section*{Preliminary Results}
-
-We have done some tests so far with a basic evolutionary algorithm \cite{github_repo} where we fixed the size of the neural network to $6$ input nodes, one hidden layer with $7$ nodes and $2$ output nodes.
-As crossover we randomly mixed the nodes with belonging weights and biases of two networks and chose the parents via the fitness proportional roulette wheel selection.
-Mutation is accomplished by adding Gaussian noise to a subset of all weights and biases.
-Furthermore we implemented elitism.
-
 The genotype-representation of the network is accomplished through python objects.
@@
 
-We implemented our own bot to act as an initial trainer for the network.
-This trainer is easier to defeat and even gives up after a certain amount of touches with the ball such that normal gameplay will be rewarded.
+\section{Standard genetic algorithm}
+
+fixed the size of the neural network to $6$ input nodes, one hidden layer with $7$ nodes and $2$ output nodes.
+
+\subsection{Crossover}
+
+As crossover we randomly mixed the nodes with belonging weights and biases of two networks and chose the parents via the fitness proportional roulette wheel selection.
+
+\subsection{Mutation}
+
+Mutation is accomplished by adding Gaussian noise to a subset of all weights and biases.
+
+\subsection{scheme}
+
+Furthermore we implemented elitism.
+
+fitness is the normalized point difference after $21$ serves.
+
+We implemented our own bot to act as an initial trainer \texttt{trainer} for the network.
+This trainer is easier to defeat.
+We made a copy of each bot where it gives up after a certain amount of touches with the ball, such that normal gameplay will be rewarded.
 The simplification gives the evolutionary algorithm the opportunity to start optimizing the neural network in a valid direction, without introducing additional factors to the fitness score.
+
+\subsection{Numerical tests}
 
@@
 \center
-\includegraphics[width=0.5\textwidth]{img/fitness.png}
-\caption{Fitness density of an example run with indicated minimum, average and maximum.}
-\label{fig:train}
+\begin{tabular}{ccc}
+\texttt{reduced} & \texttt{com\_11} & \texttt{gintonicV9} \\
+\includegraphics[width=0.3\textwidth]{img/standard_reduced.png} &
+\includegraphics[width=0.3\textwidth]{img/standard_com_11.png} &
+\includegraphics[width=0.3\textwidth]{img/standard_gintonicV9.png} \\
+\texttt{hyp014} & \texttt{Union} & \texttt{trainer} \\
+\includegraphics[width=0.3\textwidth]{img/standard_hyp014.png} &
+\includegraphics[width=0.3\textwidth]{img/standard_Union.png} &
+\includegraphics[width=0.3\textwidth]{img/standard_trainer.png}
+\end{tabular}
+\caption{Fitness density with indicated minimum, average and maximum.}
+\label{fig:standard}
 \end{figure}
 
-The results are depicted in Figure \ref{fig:train}, where the fitness is the normalized point difference after $21$ serves.
-One can see that, although we we made simplifications by weakening the opponent, the algorithm produces a network which wins all rounds.
-
-\section*{Further Experiments}
-
-We plan on testing multiple types of crossover and mutation operators, and selection schemes as seen in \cite{montana1989training}.
-Example mutations include biased and unbiased weight mutations and node mutations.
-Example crossovers are weight, node, and feature crossovers.
-
-To have a more objective representation of the capabilities of the networks, we plan on reintroducing the original easy, medium and hard difficulty bots in a curriculum-learning fashion.
-
-Finally, as time allows we could extend this work by e.g.\,making the size of the network variable or do self-training with similar techniques as in the paper of Karl Sims\cite{sims1994evolving} presented in the lecture.
+\begin{figure}[H]
+\center
+\begin{tabular}{ccc}
+\texttt{reduced\_5t} & \texttt{com\_11\_5t} & \texttt{gintonicV9\_5t} \\
+\includegraphics[width=0.3\textwidth]{img/standard_reduced_5t.png} &
+\includegraphics[width=0.3\textwidth]{img/standard_com_11_5t.png} &
+\includegraphics[width=0.3\textwidth]{img/standard_gintonicV9_5t.png} \\
+\texttt{hyp014\_5t} & \texttt{Union\_5t} & \texttt{trainer\_5t} \\
+\includegraphics[width=0.3\textwidth]{img/standard_hyp014_5t.png} &
+\includegraphics[width=0.3\textwidth]{img/standard_Union_5t.png} &
+\includegraphics[width=0.3\textwidth]{img/standard_trainer_5t.png}
+\end{tabular}
+\caption{Fitness density with indicated minimum, average and maximum.}
+\label{fig:standard_5t}
+\end{figure}
 

### 2018-12-16 project/plot_fitness.py

@@ -13,20 +13,23 @@ with open(args.filename, 'rb') as f:
 gen = len(fitness)
 
 fig = plt.figure(figsize=(10,6))
+font = {'family' : 'normal',
+        'size'   : 26}
+plt.rc('font', **font)
 
 ax = plt.axes()
 ax.plot(np.arange(gen), np.mean(fitness, axis = 1), 'k')
 ax.plot(np.arange(gen), np.max(fitness, axis = 1), 'k')
 ax.plot(np.arange(gen), np.min(fitness, axis = 1), 'k')
 
-yres = 20
-y_grid = np.linspace(0, 1, yres)
+yres = 23
+y_grid = np.linspace(0, 1.0001, yres)
 zi = [ [ np.sum(np.less_equal(y_grid[j],fitness[k])*np.less(fitness[k],y_grid[j+1])) for j in range(len(y_grid)-1)] for k in range(gen) ]
 
 zi = np.transpose(np.array(zi))
 zi = np.concatenate((zi, np.zeros([1,gen])))
 xi, yi = np.meshgrid(np.arange(0,gen)-0.5, y_grid)
-tmp = ax.pcolor(xi, yi, zi, cmap=plt.cm.Oranges)
+tmp = ax.pcolor(xi, yi, zi, cmap=plt.cm.BuGn)
 cb = plt.colorbar(tmp)
 cb.set_label('% of population')
 tmp.set_clim(vmin=0, vmax=20)
@@ -36,9 +39,6 @@ ax.set_ylabel('fitness')
 ax.set_xlim([0,gen-1.5])
 ax.set_ylim([0,1])
 
-font = {'family' : 'normal',
-        'size'   : 22}
-plt.rc('font', **font)
 plt.tight_layout()
 
 plt.show()

### 2018-12-19 project/calculate.sh
automated things a litte

@@ -1,0 +1,22 @@
+#!/bin/bash
+
+gens=1000
+bots=('reduced' 'com_11' 'gintonicV9' 'hyp014' 'Union' 'trainer')
+
+# standard
+for bot in ${bots[*]}; do
+  python3 main.py --gens $gens --bots $bot
+  python3 main.py --gens $gens --bots ${bot}_5t
+done
+
+# self_adapt_1
+for bot in ${bots[*]}; do
+  python3 main.py --self_adapt_1 --gens $gens --bots $bot
+  python3 main.py --self_adapt_1 --gens $gens --bots ${bot}_5t
+done
+
+# self_adapt_2
+for bot in ${bots[*]}; do
+  python3 main.py --self_adapt_2 --gens $gens --bots $bot
+  python3 main.py --self_adapt_2 --gens $gens --bots ${bot}_5t
+done

### 2018-12-19 project/main.py

@@ -34,15 +34,12 @@ def compute_fitness(pool, mode=None, thread_count=200):
             t.join()
 
 
-def save_fitness(fitness):
-    with open('fitness.pkl', 'wb') as f:
-        pickle.dump(fitness, f)
-
-
 def main():
     parser = argparse.ArgumentParser()
     parser.add_argument('--self_adapt_1', action='store_true')
     parser.add_argument('--self_adapt_2', action='store_true')
+    parser.add_argument('--bots', nargs='+')
+    parser.add_argument('--gens', nargs='+')
     args = parser.parse_args()
 
     # because 7 is a lucky number, our household mathemathican said so
@@ -54,24 +51,8 @@ def main():
     crossover_rate = 0.5
     mutation_p = 0.7        # mutation parameters
 
-
-#    trainers = ['reduced'];
-#    trainers = ['com_11'];
-#    trainers = ['gintonicV9'];
-#    trainers = ['hyp014'];
-#    trainers = ['Union'];
-#    trainers = ['trainer'];
-
-#    trainers = ['reduced_5t'];
-#    trainers = ['com_11_5t'];
-#    trainers = ['gintonicV9_5t'];
-#    trainers = ['hyp014_5t'];
-#    trainers = ['Union_5t'];
-    trainers = ['trainer_5t'];
-
-    gens = [200]
-
-
+    trainers = args.bots
+    gens = [ int(gen) for gen in args.gens ]
 
 #    trainers = ['trainer_5t'] # warmup
 #    gens = [20]
@@ -80,24 +61,32 @@ def main():
 
     if args.self_adapt_1:
         print('Using self adapt type 1')
+        fitness_fname = 'self_adapt_1'
         bot_class = BotSelfAdapt1
         n_elitism = 1
         def fbar(x):
             return (x + 0.2)**3
     elif args.self_adapt_2:
         print('Using self adapt type 2')
+        fitness_fname = 'self_adapt_2'
         bot_class = BotSelfAdapt2
         n_elitism = 1
         def fbar(x):
             return (x + 0.2)**3
     else:
         print('Using no self adaption')
+        fitness_fname = 'standard'
         bot_class = Bot
         mutation_rate = 0.5
         mutation_sigma = 0.01
         n_elitism = 10
         def fbar(x):
             return (x + 1)**2
+
+    for trainer in trainers:
+        fitness_fname += '_' + trainer
+    fitness_fname = '../fitnesses/' + fitness_fname
[…]

### 2018-12-19 project/plot_fitness.py

@@ -5,12 +5,16 @@ import pickle
 
 parser = argparse.ArgumentParser()
 parser.add_argument('filename', nargs = '?', default = 'fitness.pkl')
+parser.add_argument('--gens')
 args = parser.parse_args()
 
 with open(args.filename, 'rb') as f:
     fitness = pickle.load(f)
 
-gen = len(fitness)
+if args.gens is None:
+    gen = len(fitness)
+else:
+    gen = args.gens
 
 fig = plt.figure(figsize=(10,6))
 font = {'family' : 'normal',
@@ -41,4 +45,5 @@ ax.set_ylim([0,1])
 
 plt.tight_layout()
 
-plt.show()
+plt.savefig(args.filename[0:-3] + 'png')
+#plt.show()

### 2018-12-20 documentation/paper.tex
added scores of the bots competing against each other
some calculations and adding them into the documentation

@@
 
+\begin{figure}
+\begin{tabular}{c|ccccccc}
+vs. & 
+\texttt{ Axji-0-2 } &
+\texttt{ COM\_11 } &
+\texttt{ GintonicV9 } &
+\texttt{ Hypo14 } &
+\texttt{ reduced } &
+\texttt{ Union } &
+\texttt{ trainer } \\
+\hline
+\texttt{ Axji-0-2 } & - & 0 & 0 & 4 & 0 & 0 & 10 \\
+\texttt{ COM\_11 } & 21 & - & 9 & 19 & 6 & 16 & 20 \\
+\texttt{ GintonicV9 } & 21 & 12 & - & 19 & 6 & 16 & 20 \\
+\texttt{ Hypo14 } & 17 & 2 & 2 & - & 2 & 5 & 15 \\
+\texttt{ reduced } & 21 & 19 & 15 & 19 & - & 19 & 21 \\
+\texttt{ Union } & 21 & 14 & 5 & 16 & 2 & - & 20 \\
+\texttt{ trainer } & 11 & 0 & 1 & 6 & 0 & 1 & -
+\end{tabular}
+\caption{scores of bots competing against each other}
+\end{figure}
+
 \section{Standard genetic algorithm}
@@
 
+\section{Self adapt 1}
+
+\subsection{Numerical tests}
+
+\begin{figure}[H]
+\center
+\begin{tabular}{ccc}
+\texttt{reduced} & \texttt{com\_11} & \texttt{gintonicV9} \\
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_reduced.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_com_11.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_gintonicV9.png} \\
+\texttt{hyp014} & \texttt{Union} & \texttt{trainer} \\
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_hyp014.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_Union.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_trainer.png}
+\end{tabular}
+\caption{Fitness density with indicated minimum, average and maximum.}
+\label{fig:self_adapt_1}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\begin{tabular}{ccc}
+\texttt{reduced\_5t} & \texttt{com\_11\_5t} & \texttt{gintonicV9\_5t} \\
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_reduced_5t.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_com_11_5t.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_gintonicV9_5t.png} \\
+\texttt{hyp014\_5t} & \texttt{Union\_5t} & \texttt{trainer\_5t} \\
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_hyp014_5t.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_Union_5t.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_trainer_5t.png}
+\end{tabular}
+\caption{Fitness density with indicated minimum, average and maximum.}
+\label{fig:self_adapt_1_5t}
+\end{figure}
+
+\section{Self adapt 2}
+
+\subsection{Numerical tests}
+
+Parameter tuning may be needed.
+
+\begin{figure}[H]
+\center
+\begin{tabular}{ccc}
+\texttt{reduced} & \texttt{com\_11} & \texttt{gintonicV9} \\
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_reduced.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_com_11.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_gintonicV9.png} \\
+\texttt{hyp014} & \texttt{Union} & \texttt{trainer} \\
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_hyp014.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_Union.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_trainer.png}
+\end{tabular}
+\caption{Fitness density with indicated minimum, average and maximum.}
+\label{fig:self_adapt_2}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\begin{tabular}{ccc}
+\texttt{reduced\_5t} & \texttt{com\_11\_5t} & \texttt{gintonicV9\_5t} \\
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_reduced_5t.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_com_11_5t.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_gintonicV9_5t.png} \\
+\texttt{hyp014\_5t} & \texttt{Union\_5t} & \texttt{trainer\_5t} \\
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_hyp014_5t.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_Union_5t.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_trainer_5t.png}
+\end{tabular}
+\caption{Fitness density with indicated minimum, average and maximum.}
+\label{fig:self_adapt_2_5t}
+\end{figure}
+
 \bibliography{refs}{}

### 2018-12-20 project/main.py
first try for self training

@@ -59,23 +59,24 @@ def main():
 #    trainers.extend(['com_11_5t', 'gintonicV9_5t', 'hyp014_5t', 'Union_5t', 'reduced_5t'])
 #    gens.extend([15]*5)
 
+
     if args.self_adapt_1:
         print('Using self adapt type 1')
-        fitness_fname = 'self_adapt_1'
+        fname = 'self_adapt_1'
         bot_class = BotSelfAdapt1
         n_elitism = 1
         def fbar(x):
             return (x + 0.2)**3
     elif args.self_adapt_2:
         print('Using self adapt type 2')
-        fitness_fname = 'self_adapt_2'
+        fname = 'self_adapt_2'
         bot_class = BotSelfAdapt2
         n_elitism = 1
         def fbar(x):
             return (x + 0.2)**3
     else:
         print('Using no self adaption')
-        fitness_fname = 'standard'
+        fname = 'standard'
         bot_class = Bot
         mutation_rate = 0.5
         mutation_sigma = 0.01
@@ -84,15 +85,15 @@ def main():
             return (x + 1)**2
 
     for trainer in trainers:
-        fitness_fname += '_' + trainer
-    fitness_fname = '../fitnesses/' + fitness_fname
-    fitness_fname += '.pkl'
+        fname += '_' + trainer
 
 
     pool = [bot_class.random(size) for _ in range(N)]
     compute_fitness(pool, 'all')
     inverse_fitness_sort(pool)
     fitness = [np.array([bot.fitness for bot in pool])]
+
+    pool[0].write_lua('blobby-1.0_fast/data/.blobby/scripts/last_best.lua')
 
     gen = 1;
     for maxgen, trainer in zip(gens, trainers):
@@ -115,9 +116,9 @@ def main():
                 compute_fitness(pool)
             fitness_time = time.time() - fitness_time
             inverse_fitness_sort(pool)
-            pool[0].write_lua('nn_max.lua')
             pool = pool[0:N]
             fitness.append(np.array([bot.fitness for bot in pool]))
+            pool[0].write_lua('blobby-1.0_fast/data/.blobby/scripts/last_best.lua')
 
             print('Gen {}/{}; \ttime: {:2.4f}s/{:2.4f}s; \tmin = {:1.4f}; \tmean = {:1.4f}; \tmax = {:1.4f}'.format(
                 gen, np.sum(gens),
@@ -126,9 +127,9 @@ def main():
                 np.mean(fitness[-1]),
[…]

### 2018-12-20 project/bot.py
little things and a small tests before a bigger run

@@ -41,12 +41,12 @@ class Bot:
         return len(self.size)
 
     @classmethod
-    def random(cls, size):
+    def random(cls, size, trainer = 'trainer'):
         # creates a uniformly random bbot
         weights = []
         for j in range(len(size) - 1):
             weights.append(np.random.rand(size[j + 1], size[j] + 1) - 0.5)
-        bbot = cls(weights)
+        bbot = cls(weights, trainer)
 
         return bbot
 
@@ -117,15 +117,15 @@ class Bot:
 
 
 class BotSelfAdapt1(Bot):
-    def __init__(self, weights):
-        super().__init__(weights)
+    def __init__(self, weights, trainer = 'trainer'):
+        super().__init__(weights, trainer)
         self.mutation_rate = .0
         self.mutation_sigma = .0
 
     @classmethod
-    def random(cls, size):
+    def random(cls, size, trainer = 'trainer'):
         # creates a uniformly random bbot
-        bbot = super().random(size)
+        bbot = super().random(size, trainer)
 
         bbot.mutation_rate = np.random.rand() / 2 + 0.25
         bbot.mutation_sigma = 0.03 * np.random.rand()
@@ -144,13 +144,13 @@ class BotSelfAdapt1(Bot):
 
 
 class BotSelfAdapt2(Bot):
-    def __init__(self, weights):
-        super().__init__(weights)
+    def __init__(self, weights, trainer = 'trainer'):
+        super().__init__(weights, trainer)
         self.mutation_sigma = []
 
     @classmethod
-    def random(cls, size):
-        bbot = super().random(size)
+    def random(cls, size, trainer = 'trainer'):
+        bbot = super().random(size, trainer)
 
         bbot.mutation_sigma = [
             0.02 * np.random.rand(size[j + 1], size[j] + 1) for j in range(len(size) - 1)]
@@ -161,6 +161,6 @@ class BotSelfAdapt2(Bot):
         self.fitness = -np.inf
         for j in range(self.nlayers - 1):
             self.mutation_sigma[j] *= np.exp(
-                0.4 * np.random.randn(self.size[j + 1], self.size[j] + 1))
+                0.2 * np.random.randn(self.size[j + 1], self.size[j] + 1))
             self.weights[j] += self.mutation_sigma[j] * \
                 np.random.randn(self.size[j + 1], self.size[j] + 1)

### 2018-12-20 project/calculate.sh

@@ -20,3 +20,13 @@ for bot in ${bots[*]}; do
   python3 main.py --self_adapt_2 --gens $gens --bots $bot
   python3 main.py --self_adapt_2 --gens $gens --bots ${bot}_5t
 done
+
+gens=5000
+
+# self training 1
+python3 main.py --gens $gens --bots 'last_best'
+python3 main.py --self_adapt_1 --gens $gens --bots 'last_best'
+
+# self training 1
+python3 self_training_2.py --gens $gens
+python3 self_training_2.py --self_adapt_1 --gens $gens

### 2018-12-20 project/main.py

@@ -88,12 +88,13 @@ def main():
         fname += '_' + trainer
 
 
-    pool = [bot_class.random(size) for _ in range(N)]
+    pool = [bot_class.random(size, trainers[0]) for _ in range(N)]
+    pool[0].write_lua('blobby-1.0_fast/data/.blobby/scripts/last_best.lua') # it is just a random one (important if we do self training)
     compute_fitness(pool, 'all')
     inverse_fitness_sort(pool)
     fitness = [np.array([bot.fitness for bot in pool])]
+    pool[0].write_lua('blobby-1.0_fast/data/.blobby/scripts/last_best.lua') # this is really the current best
 
-    pool[0].write_lua('blobby-1.0_fast/data/.blobby/scripts/last_best.lua')
 
     gen = 1;
     for maxgen, trainer in zip(gens, trainers):
@@ -110,10 +111,7 @@ def main():
                 mutate_gaussian(offspring, mutation_p, mutation_rate, mutation_sigma)
             pool = pool[0:n_elitism] + offspring
             fitness_time = time.time()
-            if tr_gen % 5 == 0:
-                compute_fitness(pool, 'all')
-            else:
-                compute_fitness(pool)
+            compute_fitness(pool, 'all')
             fitness_time = time.time() - fitness_time
             inverse_fitness_sort(pool)
             pool = pool[0:N]

### 2018-12-20 project/self_training_2.py

@@ -1,0 +1,145 @@
+#!/bin/bash
+import argparse
+import math
+import pickle
+import threading
+import time
+from copy import deepcopy
+
+import matplotlib.pyplot as plt
+import numpy as np
+
+from bot import Bot, BotSelfAdapt1, BotSelfAdapt2
+from genetic_operators import crossover_roulette_wheel, mutate_gaussian
+
+
+def inverse_fitness_sort(pool):
+    pool.sort(key=lambda bot: -bot.fitness)
+
+
+def compute_fitness(pool, mode=None, thread_count=200):
+    # computes the fitness of the pool
+    if mode == "all":
+        subpool = pool
+    else:  # don't compute fitnesses we already know
+        subpool = [bot for bot in pool if bot.fitness == -np.inf]
+    chunks = np.array_split(subpool, math.ceil(len(subpool) / thread_count))
+
+    for chunk in chunks:
+        threads = [threading.Thread(target=bot.compute_fitness, kwargs={
+                                    'id': id}) for id, bot in enumerate(chunk)]
+        for t in threads:
+            t.start()
+        for t in threads:
+            t.join()
+
+
+def main():
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--self_adapt_1', action='store_true')
+    parser.add_argument('--self_adapt_2', action='store_true')
+    parser.add_argument('--gens')
+    args = parser.parse_args()
+
+    # because 7 is a lucky number, our household mathemathican said so
+    np.random.seed(7)
+    size = [6, 7, 2]        # size of the neuronal networks
+    N = 100                 # population size
+    n_offspring = 150
+    crossover_p = 0.9       # crossove parameters
+    crossover_rate = 0.5
+    mutation_p = 0.7        # mutation parameters
+
+    gens = int(args.gens)
+
+    if args.self_adapt_1:
+        print('Using self adapt type 1')
+        fname = 'self_adapt_1_self_training_2'
+        bot_class = BotSelfAdapt1
+        n_elitism = 1
+        def fbar(x):
+            return (x + 0.2)**3
+    elif args.self_adapt_2:
+        print('Using self adapt type 2')
+        fname = 'self_adapt_2_self_training_2'
+        bot_class = BotSelfAdapt2
+        n_elitism = 1
+        def fbar(x):
+            return (x + 0.2)**3
+    else:
[…]

### 2018-12-20 project/plot_bot_against_all.py
added function to plot the fitness of all bots against the inbuild bots

@@ -1,0 +1,56 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import glob
+import subprocess
+import os
+
+# Get templates once at import
+with open('blobby-1.0_fast/data/.blobby/config.xml', 'r') as f:
+    config_template = f.read()
+
+
+bots_x =  glob.glob("../bots/*")
+bots_x = [ os.path.basename(bot) for bot in bots_x ]
+bots_x = [ bot[0:-4] for bot in bots_x ]
+bots_y = ['reduced', 'com_11', 'gintonicV9', 'hyp014', 'Union', 'trainer']
+
+fitnesses = []
+for bot in bots_x:
+    fitnesses.append([])
+    with open('../bots/' + bot + '.lua', 'r') as src, open('blobby-1.0_fast/data/.blobby/scripts/' + bot + '.lua', 'w') as dst: dst.write(src.read())
+    for opponent in bots_y:
+        config_fname = 'config_plot.xml'
+        config_path = 'blobby-1.0_fast/data/.blobby/' + config_fname
+        config_string = config_template.replace('<var name="left_script_name" value=""/>',
+                                                     '<var name="left_script_name" value="{:s}"/>'.format(bot))
+        config_string = config_string.replace('<var name="right_script_name" value=""/>',
+                                                     '<var name="right_script_name" value="{:s}"/>'.format(opponent))
+        with open(config_path, 'w') as f:
+            f.write(config_string)
+
+        score = subprocess.run(['./src/blobby', config_fname],
+                               stdout=subprocess.PIPE,
+                               cwd='blobby-1.0_fast')
+
+        score = score.stdout.decode('utf-8').split('\n')
+        score = score[len(score) - 2].split(':')
+        score_left = int(score[1])
+        score_right = int(score[2])
+
+        fitnesses[-1].append(score_left / (score_left + score_right))
+    os.remove('blobby-1.0_fast/data/.blobby/scripts/' + bot + '.lua')
+
+
+p = []
+for j in range(len(bots_y)):
+    bottom = [ sum(s[0:j]) for s in fitnesses ]
+    cur = [ s[j] for s in fitnesses ]
+    p.append(plt.bar(range(len(bots_x)), cur, 0.5, bottom))
+
+plt.ylabel('sum over fitness')
+plt.title('fitnesses')
+plt.xticks(rotation=70)
+plt.xticks(range(len(bots_x)), bots_x)
+plt.legend(p, bots_y)
+
+plt.show()

### 2019-01-04 documentation/paper.tex

@@
 \usepackage{graphicx} % Bilder
+\usepackage{tikz, pgfplots}
+\pgfplotsset{compat=1.14}
 %}}}
@@
 %}}}
-%{{{ a todo command
+% a todo command===============================================================
+\usepackage[ruled]{algorithm2e}
 \newcounter{todocounter}
 \newcommand{\todo}[2][noisnotdefined]{
- \marginpar{\fcolorbox{black}{yellow}{\footnotesize\textbf{todo}}
- \ifthenelse{\equal{#1}{noisnotdefined}}{}{\textcolor{black}{\newline\tiny #1}} }
+ \marginpar{\fcolorbox{black}{orange}{\footnotesize\textbf{todo}}
+ \ifthenelse{\equal{#1}{noisnotdefined}}{}{\textcolor{black}{\newline\tiny #1}}}
  \textbf{\ifthenelse{\equal{#2}{.}}
-   {\fcolorbox{blue}{white}{\textcolor{blue}{$\maltese$} } }{ {\textcolor{blue}{#2} } } }
- \refstepcounter{todocounter} }
-%}}}
+   {\fcolorbox{orange}{white}{\textcolor{orange}{$\maltese$}}}{{\textcolor{orange}{#2}}}}
+ \refstepcounter{todocounter}}
+
+%===============================================================================
 
@@
 \section{Introduction}
+
+\todo{include reference \cite{Garis90geneticprogramming} (beginning of neuroevolution, cited by Sims)}
+\todo{include reference \cite{Sims94evolving3d} (Sims himself)}
 
@@
 
-\begin{figure}
+\begin{figure}[H]
 \begin{tabular}{c|ccccccc}
@@
 \end{figure}
+
+\todo{explain the table}
 
@@
 \center
+\includegraphics[width=0.4\textwidth]{img/standard.png}
+\caption{trained bots against all other}
+\label{fig:standard}
+\end{figure}
+
+\todo{coclusion: bots too hard we therefore reward gameplay by the *\_5t bots}
+
+\todo{introduce the goal of an overall good bot and explain the plot along the way (added fitnesses (score of 12 would mean winneing every point against every bot))}
+
+\begin{figure}[H]
+\center
 \begin{tabular}{ccc}
@@
 
+\begin{figure}[H]
+\center
+\includegraphics[width=0.4\textwidth]{img/standard_5t.png}
+\caption{trained bots against all other}
+\label{fig:standard_5t}
+\end{figure}
+
+\todo{conclusion: better, and not only for 5t bots (algorithm has something to grab on)}
+
+\todo{overall: not very pretty graphs -> need self adaption}
+
 \section{Self adapt 1}
+
+\todo{explain what we are doing}
 
@@
 \center
+\includegraphics[width=0.4\textwidth]{img/self_adapt_1.png}
+\caption{trained bots against all other}
+\label{fig:self_adapt_1}
+\end{figure}
+
+\todo{we can see that the algorithm is converging}
+\todo{because trainer is beatable, the nn has something to grab on and performs way better thatn the others}
+
+\begin{figure}[H]
+\center
 \begin{tabular}{ccc}
@@
 
+\begin{figure}[H]
+\center
+\includegraphics[width=0.4\textwidth]{img/self_adapt_2_5t.png}
+\caption{trained bots against all other}
+\label{fig:self_adapt_2_5t}
+\end{figure}
+
+\todo{now it's pretty}
+\todo{trainer\_5t bad because it is very easy to beat and therefore we can get loast in a local optimum real quick}
+
 \section{Self adapt 2}
@@
 \subsection{Numerical tests}
-
-Parameter tuning may be needed.
 
@@
 \center
+\includegraphics[width=0.4\textwidth]{img/self_adapt_2.png}
+\caption{trained bots against all other}
+\label{fig:self_adapt_2}
+\end{figure}
+
+\todo{similar to section before without 5t}
+
+\begin{figure}[H]
+\center
 \begin{tabular}{ccc}
@@
 
+\begin{figure}[H]
+\center
+\includegraphics[width=0.4\textwidth]{img/self_adapt_2_5t.png}
+\caption{trained bots against all other}
+\label{fig:self_adapt_2_5t}
+\end{figure}
+
+\todo{not very predictable behaviour}
+\todo{works good for a few examples}
+\todo{remember prof has never seen this stuff working and will be happy if it has disadvantages}
+
+\section{Changing opponent}
+
+200 gen \texttt{gintonicV9\_5t}, 200 gen \texttt{com\_11\_5t}, 200 gen \texttt{gintonicV9}, 200 gen \texttt{com\_11}.
+\todo{argument with the previous plots that this is a good idea}
+\todo{other ideas?}
+
+\begin{figure}[H]
+\center
+\begin{tabular}{ccc}
+standard & self adapt 1 & self adapt 2 \\
+\includegraphics[width=0.3\textwidth]{img/standard_gintonicV9_5t_com_11_5t_gintonicV9_com_11.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_gintonicV9_5t_com_11_5t_gintonicV9_com_11.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_2_gintonicV9_5t_com_11_5t_gintonicV9_com_11.png} \\
+\end{tabular}
+\caption{Fitness density with indicated minimum, average and maximum.}
+\label{fig:changing opponent}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\includegraphics[width=0.4\textwidth]{img/change.png}
+\caption{trained bots against all other}
+\label{fig:change}
+\end{figure}
+
+\todo{only self adapt 1 works}
+\todo{we could do more with more generations but loose the comparebility to the previous section}
+
+\section{Self-training}
+
+\subsection{last best}
+
+\begin{figure}[H]
+\center
+\includegraphics[width=0.155555\textwidth]{img/scheme_self_training_1.png}
+\caption{Scheme for self training variant 1}
+\label{fig:scheme_self_training_1}
+\end{figure}
+
+Fitness evaluated with the best bot of the last generation.
+
+\begin{figure}[H]
+\center
+\begin{tabular}{cc}
+standard & self adapt 1\\
+\includegraphics[width=0.3\textwidth]{img/standard_last_best.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_last_best.png}
+\end{tabular}
+\caption{Fitness density with indicated minimum, average and maximum.}
+\label{fig:last_best}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\includegraphics[width=0.3\textwidth]{img/self_training_1.png}
+\caption{trained bots against all other}
+\label{fig:self_training_1}
+\end{figure}
+
+\todo{risk to end up in local optimum like self adapt 1?}
+
+\subsection{two species}
+
+\begin{figure}[H]
+\center
+\includegraphics[width=0.155555\textwidth]{img/scheme_self_training_1.png}
+\caption{Scheme for self training variant 1}
+\label{fig:scheme_self_training_2}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\begin{tabular}{cc}
+left side & right side \\
+\includegraphics[width=0.3\textwidth]{img/standard_self_training_2_1.png} &
+\includegraphics[width=0.3\textwidth]{img/standard_self_training_2_2.png} \\
+\end{tabular}
+\caption{Fitness density with indicated minimum, average and maximum standard mutation.}
+\label{fig:standard_self_training_2}
+\end{figure}
+
+\todo{too much diversity}
+
+\begin{figure}[H]
+\center
+\begin{tabular}{cc}
+left side & right side \\
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_self_training_2_1_oldrun.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_self_training_2_2_oldrun.png}
+\end{tabular}
+\caption{Fitness density with indicated minimum, average and maximum with self adapt 1.}
+\label{fig:self_adapt_1_self_training_2_oldrun}
+\end{figure}
+
+\todo{sadly they do not change too often}
+\todo{there is still improvement even though it can not be seen in the plots because they improve themselves}
+
+\begin{figure}[H]
+\center
+\begin{tabular}{cc}
+left side & right side \\
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_self_training_2_1.png} &
+\includegraphics[width=0.3\textwidth]{img/self_adapt_1_self_training_2_2.png} \\
+\end{tabular}
+\caption{Fitness density with indicated minimum, average and maximum with self adapt 1 and more generations.}
+\label{fig:self_adapt1_self_training_2}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\includegraphics[width=0.4\textwidth]{img/self_training_2.png}
+\caption{trained bots against all other}
+\label{fig:self_training_2}
+\end{figure}
+
+\todo{we got a pretty awesome bot there which gets beaten only by reduced but never trained against it}
+
+\section{Discussion and Future Ideas}
+
+\todo{everything we have not done}
+\todo{train the self trained bot with reduced so we maybe get a perfect score of 12}
+\todo{we have not used too many inputs I still coudn't beat the bot which trained itself for four hours}
+\todo{acknowledgement to the compute server}
+
 \bibliography{refs}{}

### 2019-01-05 documentation/paper.tex
Work-work

@@
 
-\todo{include reference \cite{Garis90geneticprogramming} (beginning of neuroevolution, cited by Sims)}
-\todo{include reference \cite{Sims94evolving3d} (Sims himself)}
+Undoubtedly, the neural network architecture has been the star of the machine learning resurgence of the past few years.
+It's provable quality of being an universal function approximator makes it a sensible choice for various applications from control to .
+The main algorithm used for training, backpropagation however is subject to an array of criticism from various authors \cite{some_papers_about_backprop}, and there are various attempts at improving or replacing it \cite{that_paper_where_ga_beats_backprop}.
+\todo{Add a sentence of context for \cite{Sims94evolving3d} and \cite{Garis90geneticprogramming}}
 
-We want to apply a genetic algorithm to a neural network which is known as neuroevolution.
-A big advantage of this scheme is that one only needs an indicator how good the neural network performs given a certain task.
+In this work we explore some properties of neuroevolution, the technique of applying a genetic algorithm to a neural network \cite{Garis90geneticprogramming} \cite{Sims94evolving3d}.
+A big advantage of this scheme is that for a series of different inputs a single scalar indicator of performance is sufficient in order to optimize a neural network for a certain task.
 This can for instance be given by the outcome of a game.
-Our game of choice was Blobby Volley 2 which is a simple two-dimensional volleyball game with open source written in C++ with lua support, see Figure \ref{fig:screenshot}.
+Our game of choice is Blobby Volley 2 which is an open source, simple two-dimensional volleyball game written in C++ with lua support. See Figure \ref{fig:screenshot}.
 
@@
 
+\section{Training environment}
+
 A scene from the game could be described by the $x$- and $y$-coordinates of the players and the ball, their corresponding velocities and the current number of ball contacts.
-These arguments or a subset of them could then be used as input for the neural network.
-The output would then be the players movement which only consists of going right, left or jumping.
-Assuming the layers of the neural network have $n_1,\dots,n_L$ nodes we then have to optimize \[N = \sum_{l=1}^{L-1} (n_l+1) \, n_{l+1}\] number of variables if we fix the size of the network.
+These arguments or a subset of them can then be used as input for the neural network.
+The output should correspond to the possible movement options of the player, which consists of going right, going left, or jumping.
+Assuming the layers of the neural network have $n_1,\dots,n_L$ nodes we then have to optimize \(N = \sum_{l=1}^{L-1} (n_l+1) \, n_{l+1}\) number of variables.
 
-The genotype-representation of the network is accomplished through python objects.
-This will allow a straightforward implementation of a wide variety of genetic operators, while having next to no impact on the speed of training.
-As the evaluation of the population is time-consuming, we do this in a thread-parallel fashion, allowing us to exploit multiple high core-count machines at our disposal.
+The game comes bundled with a number of bots with different styles, that are designed to pose a challenge to a wide range of human opponents.
+In order to define a clean ranking between them, we recorded the score of all possible lineups.
+The results can be seen in table no.
+\ref{tab:bot_vs_bot}.
+\todo{explain the table}
 
@@
 vs. &
-\texttt{ Axji-0-2 } &
-\texttt{ COM\_11 } &
-\texttt{ GintonicV9 } &
-\texttt{ Hypo14 } &
-\texttt{ reduced } &
-\texttt{ Union } &
-\texttt{ trainer } \\
+\texttt{Axji-0-2} &
+\texttt{COM\_11} &
+\texttt{GintonicV9} &
+\texttt{Hypo14} &
+\texttt{reduced} &
+\texttt{Union} &
+\texttt{trainer} \\
 \hline
-\texttt{ Axji-0-2 } & - & 0 & 0 & 4 & 0 & 0 & 10 \\
-\texttt{ COM\_11 } & 21 & - & 9 & 19 & 6 & 16 & 20 \\
-\texttt{ GintonicV9 } & 21 & 12 & - & 19 & 6 & 16 & 20 \\
-\texttt{ Hypo14 } & 17 & 2 & 2 & - & 2 & 5 & 15 \\
-\texttt{ reduced } & 21 & 19 & 15 & 19 & - & 19 & 21 \\
-\texttt{ Union } & 21 & 14 & 5 & 16 & 2 & - & 20 \\
-\texttt{ trainer } & 11 & 0 & 1 & 6 & 0 & 1 & -
+\texttt{Axji-0-2} & - & 0 & 0 & 4 & 0 & 0 & 10 \\
+\texttt{COM\_11} & 21 & - & 9 & 19 & 6 & 16 & 20 \\
+\texttt{GintonicV9} & 21 & 12 & - & 19 & 6 & 16 & 20 \\
+\texttt{Hypo14} & 17 & 2 & 2 & - & 2 & 5 & 15 \\
+\texttt{reduced} & 21 & 19 & 15 & 19 & - & 19 & 21 \\
+\texttt{Union} & 21 & 14 & 5 & 16 & 2 & - & 20 \\
+\texttt{trainer} & 11 & 0 & 1 & 6 & 0 & 1 & -
 \end{tabular}
-\caption{scores of bots competing against each other}
+\label{tab:bot_vs_bot}
+\caption{Scores of bots competing against each other}
 \end{figure}
 
-\todo{explain the table}
+\section{Implementation}
+The genotype-representation of the network is accomplished through python objects.
+This allows a straightforward implementation of a wide variety of genetic operators.
+Since our algorithm is constrained overwhelmingly by evaluation time, these inefficient but easy to modify structures do not impact performance.
+As the evaluation of the population is time-consuming, we do this in a thread-parallel fashion, allowing us to exploit multiple high core-count machines at our disposal.
 
@@
 
-fixed the size of the neural network to $6$ input nodes, one hidden layer with $7$ nodes and $2$ output nodes.
+\subsection{Network architecture}
+
+For the whole experiment, we fixed the size of the neural network to $6$ input nodes, one hidden layer with $7$ nodes and $2$ output nodes.
+As can be seen later, this proved sufficient for a surprisingly high performance for the network.
+The input nodes got fed with the x and y positions of the player, it's relative position compared to the ball, and the velocity vector of the ball.
+The output nodes consisted of two neurons with tanh activations, one responsible for going left or right in the next timestep (if the output was below or above of threshold values $t_l = 0.49, t_r = 0.51$) and the other for jumping (when above threshold $t_j = 0.7$).
 
@@
 
-\subsection{scheme}
+\subsection{Scheme}
 
@@
 \includegraphics[width=0.4\textwidth]{img/standard_5t.png}
-\caption{trained bots against all other}
-\label{fig:standard_5t}
+\caption{Trained bots against all other}
+\label{fig:standard_5t_aa}
 \end{figure}
@@
 \caption{trained bots against all other}
-\label{fig:self_adapt_1}
+\label{fig:self_adapt_1_aa}
 \end{figure}
@@
 \caption{trained bots against all other}
-\label{fig:self_adapt_2_5t}
+\label{fig:self_adapt_2_5t_aa}
 \end{figure}
@@
 \caption{trained bots against all other}
-\label{fig:self_adapt_2}
+\label{fig:self_adapt_2_aa}
 \end{figure}

### 2019-01-05 documentation/paper.tex
More work-work

@@
 
+We implemented our own bot to act as an initial trainer \texttt{trainer} for the network.
+This trainer is easier to defeat.
+We made a copy of each bot where it gives up after a certain amount of touches with the ball, such that normal gameplay will be rewarded.
+The simplification gives the evolutionary algorithm the opportunity to start optimizing the neural network in a valid direction, without introducing additional factors to the fitness score.
+
 \section{Implementation}
@@
 
-\subsection{Crossover}
+\subsection{Scheme}
+The algorithm worked with a population size of 100, with 150 offspring after crossover.
+Our survivor selection strategy was elitism, selecting from the pool of children plus the 10 best parents from the last generation.
 
-As crossover we randomly mixed the nodes with belonging weights and biases of two networks and chose the parents via the fitness proportional roulette wheel selection.
+The fitness function was the normalized point difference after $21$ serves.
 
 \subsection{Mutation}
+Mutation is accomplished by adding Gaussian noise to a subset of all weights and biases. \todo{Default values, rate of change}
 
-Mutation is accomplished by adding Gaussian noise to a subset of all weights and biases.
-
-\subsection{Scheme}
-
-Furthermore we implemented elitism.
-
-fitness is the normalized point difference after $21$ serves.
-
-We implemented our own bot to act as an initial trainer \texttt{trainer} for the network.
-This trainer is easier to defeat.
-We made a copy of each bot where it gives up after a certain amount of touches with the ball, such that normal gameplay will be rewarded.
-The simplification gives the evolutionary algorithm the opportunity to start optimizing the neural network in a valid direction, without introducing additional factors to the fitness score.
+\subsection{Crossover}
+As crossover we randomly mixed the nodes with belonging weights and biases of two networks and chose the parents via the fitness proportional roulette wheel selection. \todo{cite roulett wheel sel.}
 
@@
 \end{tabular}
-\caption{Fitness density with indicated minimum, average and maximum.}
+\caption{Fitness density with indicated minimum, average and maximum while training against different bots. \todo{is this true?}}
 \label{fig:standard}

### 2019-01-05 documentation/paper.tex
minor workwork

@@
 For the whole experiment, we fixed the size of the neural network to $6$ input nodes, one hidden layer with $7$ nodes and $2$ output nodes.
-As can be seen later, this proved sufficient for a surprisingly high performance for the network.
+As can be seen later, this proved sufficient for a surprisingly high performance by the network.
 The input nodes got fed with the x and y positions of the player, it's relative position compared to the ball, and the velocity vector of the ball.
-The output nodes consisted of two neurons with tanh activations, one responsible for going left or right in the next timestep (if the output was below or above of threshold values $t_l = 0.49, t_r = 0.51$) and the other for jumping (when above threshold $t_j = 0.7$).
+The output nodes consisted of two neurons with tanh activations, one responsible for going left or right in the next timestep (if the output was below or above of threshold values $t_l = 0.49, t_r = 0.51$) and the other for jumping (above threshold $t_j = 0.7$).
 
@@
 The algorithm worked with a population size of 100, with 150 offspring after crossover.
-Our survivor selection strategy was elitism, selecting from the pool of children plus the 10 best parents from the last generation.
+Our survivor selection strategy was elitism, selecting from the pool of children, plus the 10 best parents from the last generation.
 
@@
 \includegraphics[width=0.4\textwidth]{img/standard.png}
-\caption{trained bots against all other}
+\caption{trained bots against all other \todo{explain this pls}}
 \label{fig:standard}

### 2019-01-07 documentation/paper.tex
minor workwork
Holy hell it's late

@@
 \usepackage{float} % use H in figure placement
-\pagenumbering{gobble}
+% \pagenumbering{gobble}
 
@@
 
-\date{}
-\title{todo: think about good title}
+\date{\today}
+\title{Really Good Title}
+\subtitle{Genetic Algorithms and Evolutionary Computing Project}
 \author{Felix Bartel and D\'avid Kerekes}
@@
 \maketitle
+\todo{think about good title}
 
@@
 Undoubtedly, the neural network architecture has been the star of the machine learning resurgence of the past few years.
-It's provable quality of being an universal function approximator makes it a sensible choice for various applications from control to .
+It's provable quality of being an universal function approximator makes it a sensible choice for various applications.
 The main algorithm used for training, backpropagation however is subject to an array of criticism from various authors \cite{some_papers_about_backprop}, and there are various attempts at improving or replacing it \cite{that_paper_where_ga_beats_backprop}.
@@
 
-In this work we explore some properties of neuroevolution, the technique of applying a genetic algorithm to a neural network \cite{Garis90geneticprogramming} \cite{Sims94evolving3d}.
+In this project we explore some properties of neuroevolution, the technique of applying a genetic algorithm to a neural network \cite{Garis90geneticprogramming} \cite{Sims94evolving3d}.
 A big advantage of this scheme is that for a series of different inputs a single scalar indicator of performance is sufficient in order to optimize a neural network for a certain task.
@@
 As the evaluation of the population is time-consuming, we do this in a thread-parallel fashion, allowing us to exploit multiple high core-count machines at our disposal.
+In order to achieve this, we modified the original game by replacing the graphics and sound modules with dummies.
 
-\section{Standard genetic algorithm}
+\section{Initial genetic algorithm}
+Here we describe our starting point for the genetic algorithm. All subsequent versions are based off of this variant, with not mentioned details staying the same.
 
@@
 The algorithm worked with a population size of 100, with 150 offspring after crossover.
-Our survivor selection strategy was elitism, selecting from the pool of children, plus the 10 best parents from the last generation.
+Our survivor selection strategy was elitism, selecting from the pool of children plus the 10 best parents from the last generation.
 
@@
 \subsection{Mutation}
-Mutation is accomplished by adding Gaussian noise to a subset of all weights and biases. \todo{Default values, rate of change}
+Mutation is accomplished by adding Gaussian noise to a subset of all weights and biases.
+We selected an individual for mutation with a probability of $0.7$, which then had a per-neuron mutation rate of $0.5$, and a sigma of $0.01$.
 
 \subsection{Crossover}
-As crossover we randomly mixed the nodes with belonging weights and biases of two networks and chose the parents via the fitness proportional roulette wheel selection. \todo{cite roulett wheel sel.}
+As crossover we randomly mixed the nodes with belonging weights and biases of two networks and chose the parents via the fitness proportional roulette wheel selection.
 
 \subsection{Numerical tests}
+We trained six networks, each against a different bot.
+We test the generalization of both the training method and the best specimens by evaluating all the networks against all the bots.
+A score for each network is calculated by summing up it's fitness value for all cases (see Fig.
+\ref{fig:standard_c}), so a network consistently beating all opponents would receive a score of 12.
+We expect that a good training method would produce networks that all beat their respective trainers and also fare well against other opponents.
+The chances of happening upon a network that can exploit a trainer by "accident" (but is susceptible to abuse from a human player for example) decrease exponentially by evaluating on multiple opponents.
 
@@
 \end{tabular}
-\caption{Fitness density with indicated minimum, average and maximum while training against different bots. \todo{is this true?}}
+\caption{Fitness density with indicated minimum, average and maximum while training against different bots.}
 \label{fig:standard}
@@
 \includegraphics[width=0.4\textwidth]{img/standard.png}
-\caption{trained bots against all other \todo{explain this pls}}
-\label{fig:standard}
+\caption{Cumulative sum of the trained neural networks against all the bots.}
+\label{fig:standard_c}
 \end{figure}
 
-\todo{coclusion: bots too hard we therefore reward gameplay by the *\_5t bots}
-
-\todo{introduce the goal of an overall good bot and explain the plot along the way (added fitnesses (score of 12 would mean winneing every point against every bot))}
+We can see that the algorithm achieves reasonable scores against the two easiest opponents, \texttt{trainer} and \texttt{hyp014}.
+Since we were not able to get significantly better results by hyperparameter-tweaking, we theorized that the discrepancy might come from the trainers being too hard to beat by any initial network mutation, therefore providing the same zero fitness for any single useful mutation.
+Therefore we introduced a weaker, "benevolent" set of trainers that give up after scoring five points by not returning the ball anymore.
+This lets a specimen that can pass initial serves back receive a positive fitness score.
+These bots are marked by appending the \texttt{\_5t} suffix to their name.
 
@@
 \includegraphics[width=0.4\textwidth]{img/standard_5t.png}
-\caption{Trained bots against all other}
+\caption{Trained networks against all other bots.}
 \label{fig:standard_5t_aa}
@@
 
-\todo{conclusion: better, and not only for 5t bots (algorithm has something to grab on)}
+As Fig.
+\ref{fig:standard_5t} and \ref{fig:standard_5t_aa} show, the networks evolved against the benevolent trainers give significantly better performance against the whole batch of bots, reaching twice the maximum cumulative score.
 
-\todo{overall: not very pretty graphs -> need self adaption}
+\section{Per-individual self adaptation}
 
-\section{Self adapt 1}
+\todo{overall: not very pretty graphs <<-- Why do we decided to do self adaptation?}
 
-\todo{explain what we are doing}
+We first implemented a method similar to (4) from \cite{self_adapt}, giving each individual a mutation rate and a sigma value.
+These were mutated when the individual was selected for mutation.
+Their initialization was also random, in the range of [0.25, 0.75] for the mutation rate and [0, 0.03] for the sigma values.
 
@@
 \includegraphics[width=0.4\textwidth]{img/self_adapt_1.png}
-\caption{trained bots against all other}
+\caption{Trained networks against all other bots.}
 \label{fig:self_adapt_1_aa}
 \end{figure}
-
-\todo{we can see that the algorithm is converging}
-\todo{because trainer is beatable, the nn has something to grab on and performs way better thatn the others}
 
@@
 \includegraphics[width=0.4\textwidth]{img/self_adapt_2_5t.png}
-\caption{trained bots against all other}
-\label{fig:self_adapt_2_5t_aa}
+\caption{Trained networks against all other bots.}
+\label{fig:self_adapt_1_5t_aa}
 \end{figure}
 
-\todo{now it's pretty}
-\todo{trainer\_5t bad because it is very easy to beat and therefore we can get loast in a local optimum real quick}
+We can see in figures \ref{self_adapt_1_5t} and \ref{self_adapt_1_5t_aa} that the results improve significantly in both cases.
+Self adaptation causes the networks trained on \texttt{_5t} opponents to get consistent results across the board.
 
-\section{Self adapt 2}
+\todo{trainer\_5t does not coincide on the two graphs, what happened? We get a good result against trainer_5t according to the bottom graph}
+
+\section{Per-parameter self adaptation}
+
+We also explored a method similar to (5) in \cite{self_adapt}, where each parameter gets it's own mutation sigma parameter.
+This produces similar results to the harsh teachers with per-individual self adaptation
 
@@
 \includegraphics[width=0.4\textwidth]{img/self_adapt_2.png}
-\caption{trained bots against all other}
+\caption{Trained networks against all other bots.}
 \label{fig:self_adapt_2_aa}
@@
 \includegraphics[width=0.4\textwidth]{img/self_adapt_2_5t.png}
-\caption{trained bots against all other}
+\caption{Trained networks against all other bots.}
 \label{fig:self_adapt_2_5t_aa}
@@
 \includegraphics[width=0.4\textwidth]{img/change.png}
-\caption{trained bots against all other}
+\caption{Trained networks against all other bots.}
 \label{fig:change}
@@
 \includegraphics[width=0.3\textwidth]{img/self_training_1.png}
-\caption{trained bots against all other}
+\caption{Trained networks against all other bots.}
 \label{fig:self_training_1}
@@
 \includegraphics[width=0.4\textwidth]{img/self_training_2.png}
-\caption{trained bots against all other}
+\caption{Trained networks against all other bots.}
 \label{fig:self_training_2}
@@
 \section{Discussion and Future Ideas}
+
+\todo{rationalize not messing around with more mutation and crossover types as we promised}
 

### 2019-01-07 project/genetic_operators.py

@@ -56,6 +56,7 @@ def mutate_gaussian(bots, mutation_p, *args):
             if np.random.rand() < mutation_p:
                 bots[n].mutate_gaussian(rate, sigma)
     else:
+        # Self-adapt variants don't have rate and sigma params
         for n in range(len(bots)):
             if np.random.rand() < mutation_p:
                 bots[n].mutate_gaussian()

### 2019-01-07 project/main.py

@@ -61,14 +61,14 @@ def main():
 
 
     if args.self_adapt_1:
-        print('Using self adapt type 1')
+        print('Using self adapt type 1') # Per-individual self adapt
         fname = 'self_adapt_1'
         bot_class = BotSelfAdapt1
         n_elitism = 1
         def fbar(x):
             return (x + 0.2)**3
     elif args.self_adapt_2:
-        print('Using self adapt type 2')
+        print('Using self adapt type 2') # Per-param self adapt
         fname = 'self_adapt_2'
         bot_class = BotSelfAdapt2
         n_elitism = 1

### 2019-01-07 documentation/paper.tex
Curriculum learning

@@
 \includegraphics[width=0.4\textwidth]{img/standard_5t.png}
-\caption{Trained networks against all other bots.}
+\caption{Best individuals against all bots.}
 \label{fig:standard_5t_aa}
@@
 \includegraphics[width=0.4\textwidth]{img/self_adapt_1.png}
-\caption{Trained networks against all other bots.}
+\caption{Best individuals against all bots.}
 \label{fig:self_adapt_1_aa}
@@
 \includegraphics[width=0.4\textwidth]{img/self_adapt_2_5t.png}
-\caption{Trained networks against all other bots.}
+\caption{Best individuals against all bots.}
 \label{fig:self_adapt_1_5t_aa}
@@
 
-We can see in figures \ref{self_adapt_1_5t} and \ref{self_adapt_1_5t_aa} that the results improve significantly in both cases.
-Self adaptation causes the networks trained on \texttt{_5t} opponents to get consistent results across the board.
+We can see in figures \ref{fig:self_adapt_1_5t} and \ref{fig:self_adapt_1_5t_aa} that the results improve significantly in both cases.
+Self adaptation causes the networks trained on \texttt{\_5t} opponents to get consistent results across the board.
 
-\todo{trainer\_5t does not coincide on the two graphs, what happened? We get a good result against trainer_5t according to the bottom graph}
+The surprisingly low performance against the easiest bot, \texttt{trainer\_5t} might be explained by the bot being too easy.
+The high initial fitness might encourage the survival of individuals with a low mutation rate, causing the algorithm to get stuck in a local optimum.
+
+\todo{trainer5t does not coincide on the two graphs, what happened? We get a good result against trainer5t according to the bottom graph}
 
@@
 We also explored a method similar to (5) in \cite{self_adapt}, where each parameter gets it's own mutation sigma parameter.
-This produces similar results to the harsh teachers with per-individual self adaptation
+This produces similar results to per-individual self adaptation with harsh teachers, but with benevolent teachers fails to reproduce the consistency of the other method.
 
@@
 \includegraphics[width=0.4\textwidth]{img/self_adapt_2.png}
-\caption{Trained networks against all other bots.}
+\caption{Best individuals against all bots.}
 \label{fig:self_adapt_2_aa}
 \end{figure}
-
-\todo{similar to section before without 5t}
 
@@
 \includegraphics[width=0.4\textwidth]{img/self_adapt_2_5t.png}
-\caption{Trained networks against all other bots.}
+\caption{Best individuals against all bots.}
 \label{fig:self_adapt_2_5t_aa}
@@
 
-\todo{not very predictable behaviour}
-\todo{works good for a few examples}
-\todo{remember prof has never seen this stuff working and will be happy if it has disadvantages}
+\section{Curriculum learning}
 
-\section{Changing opponent}
+We implemented a limited set of curriculum learning, where the training happens against increasingly harder opponents \texttt{gintonicV9\_5t}, \texttt{com\_11\_5t}, \texttt{gintonicV9}, and \texttt{com\_11} for 200 generations each.
+These bots previously worked as solid teachers.
+As can be seen on \ref{fig:change}, only the Per-individual self adaptation seems to work, there's no transfer of knowledge between the different bots for the other methods.
 
-200 gen \texttt{gintonicV9\_5t}, 200 gen \texttt{com\_11\_5t}, 200 gen \texttt{gintonicV9}, 200 gen \texttt{com\_11}.
-\todo{argument with the previous plots that this is a good idea}
-\todo{other ideas?}
+It's possible that with increasing the number of generations this algorithm would produce some advantages over training with just one teacher.
+Sadly, this would have to repeat all the previous experiments with a higher generation count for comparability's sake.
 
@@
 \begin{tabular}{ccc}
-standard & self adapt 1 & self adapt 2 \\
+Standard & Per-individual s. a. & Per-parameter s. a. \\
 \includegraphics[width=0.3\textwidth]{img/standard_gintonicV9_5t_com_11_5t_gintonicV9_com_11.png} &
@@
 \includegraphics[width=0.4\textwidth]{img/change.png}
-\caption{Trained networks against all other bots.}
+\caption{Best individuals against all bots.}
 \label{fig:change}
 \end{figure}
-
-\todo{only self adapt 1 works}
-\todo{we could do more with more generations but loose the comparebility to the previous section}
 
@@
 \includegraphics[width=0.3\textwidth]{img/self_training_1.png}
-\caption{Trained networks against all other bots.}
+\caption{Best individuals against all bots.}
 \label{fig:self_training_1}
@@
 \includegraphics[width=0.4\textwidth]{img/self_training_2.png}
-\caption{Trained networks against all other bots.}
+\caption{Best individuals against all bots.}
 \label{fig:self_training_2}

### 2019-01-07 documentation/paper.tex
All vs best

@@
 \section{Per-individual self adaptation}
+\label{sec:perind}
 
@@
 
-We first implemented a method similar to (4) from \cite{self_adapt}, giving each individual a mutation rate and a sigma value.
+We first implemented a method similar to (1) and (4) from \cite{self_adapt}, giving each individual a mutation rate and a sigma value.
 These were mutated when the individual was selected for mutation.
@@
 
-We also explored a method similar to (5) in \cite{self_adapt}, where each parameter gets it's own mutation sigma parameter.
+We also explored a method similar to (2) and (5) in \cite{self_adapt}, where each parameter gets it's own mutation sigma parameter.
 This produces similar results to per-individual self adaptation with harsh teachers, but with benevolent teachers fails to reproduce the consistency of the other method.
@@
 
-It's possible that with increasing the number of generations this algorithm would produce some advantages over training with just one teacher.
-Sadly, this would have to repeat all the previous experiments with a higher generation count for comparability's sake.
+It's possible that with increasing the number of generations this algorithm would produce some advantages over training with just one teacher, but we couldn't finish the longer experiment due to time constraints.
 
@@
 \section{Self-training}
+Implementing self-training based on \cite{Sims94evolving3d}.
 
-\subsection{last best}
+\subsection{All vs best}
+We ran a version of the algorithm where the fitness function was evaluated on the best specimen of the last generation.
 
@@
 \includegraphics[width=0.155555\textwidth]{img/scheme_self_training_1.png}
-\caption{Scheme for self training variant 1}
+\caption{Scheme for all vs best training; source: \cite{Sims94evolving3d}.}
 \label{fig:scheme_self_training_1}
 \end{figure}
-
-Fitness evaluated with the best bot of the last generation.
 
@@
 
-\todo{risk to end up in local optimum like self adapt 1?}
+Since we start out with evaluating against easy opponents, it's a possibility for the algorithm to get stuck in a local optimum like it did with per-individual self adaptation.
+While the standard method works nice, (see figures \ref{fig:last_best} and \ref{fig:self_training_1}) the self-adaption variant fails to produce any meaningful result despite the high fitness value, thus supporting our theory from section \ref{sec:perind}.
 
-\subsection{two species}
+\subsection{Two-species all vs best}
+
+This implementation of the idea from \cite{Sims94evolving3d} uses two species that are always evaluated against the other.
+
+\todo{Did we have two 100 sized pops here?}
 
@@
 \center
-\includegraphics[width=0.155555\textwidth]{img/scheme_self_training_1.png}
+\includegraphics[width=0.155555\textwidth]{img/scheme_self_training_2.png}
 \caption{Scheme for self training variant 1}

### 2019-01-07 documentation/paper.tex
Some stuff for two-species all vs best

@@
 \begin{tabular}{ccc}
-Standard & Per-individual s. a. & Per-parameter s. a. \\
+standard & per-individual s. a. & per-parameter s. a. \\
 \includegraphics[width=0.3\textwidth]{img/standard_gintonicV9_5t_com_11_5t_gintonicV9_com_11.png} &
@@
 
-This implementation of the idea from \cite{Sims94evolving3d} uses two species that are always evaluated against the other.
+This implementation of the idea from \cite{Sims94evolving3d} uses two species that are always evaluated against the other. We evaluated
 
@@
 \includegraphics[width=0.155555\textwidth]{img/scheme_self_training_2.png}
-\caption{Scheme for self training variant 1}
+\caption{Scheme for two-species all vs best self training; source: \cite{Sims94evolving3d}.}
 \label{fig:scheme_self_training_2}
@@
 \begin{tabular}{cc}
-left side & right side \\
+standard left side & standard right side \\
 \includegraphics[width=0.3\textwidth]{img/standard_self_training_2_1.png} &
 \includegraphics[width=0.3\textwidth]{img/standard_self_training_2_2.png} \\
-\end{tabular}
-\caption{Fitness density with indicated minimum, average and maximum standard mutation.}
-\label{fig:standard_self_training_2}
-\end{figure}
-
-\todo{too much diversity}
-
-\begin{figure}[H]
-\center
-\begin{tabular}{cc}
-left side & right side \\
+self-adapt left side & self-adapt right side \\
 \includegraphics[width=0.3\textwidth]{img/self_adapt_1_self_training_2_1_oldrun.png} &
@@
 \end{tabular}
-\caption{Fitness density with indicated minimum, average and maximum with self adapt 1.}
+\caption{Fitness density two-species all vs best method with indicated minimum, average and maximum.}
 \label{fig:self_adapt_1_self_training_2_oldrun}
 \end{figure}
+
+In this scenario \ref{fig:self_adapt_1_self_training_2_oldrun}
 

### 2019-01-07 documentation/paper.tex
Two-species all vs best

@@
 
-In this scenario \ref{fig:self_adapt_1_self_training_2_oldrun}
+\todo{too much diversity <<-- what does this mean?}
 
-\todo{sadly they do not change too often}
-\todo{there is still improvement even though it can not be seen in the plots because they improve themselves}
+In this scenario \ref{fig:self_adapt_1_self_training_2_oldrun} can be seen battling for being the better species in the beginning, but then this seems to converge.
+A quick manual check confirms however, that they are both still actively learning.
+When left alone to train for 8000 generations (figure \ref{fig:self_adapt1_self_training_2}) we get the best results so far (figure \ref{fig:self_training_2}).
+The only bot capable of matching this network is the strongest \texttt{reduced} bot, but it never trained against any of the built-in bots directly.
+(In fact, this network was too strong even for us humans to beat in the game.)
 
@@
 
-\todo{we got a pretty awesome bot there which gets beaten only by reduced but never trained against it}
-
 \section{Discussion and Future Ideas}

### 2019-01-07 documentation/paper.tex
Rework introduction, add cites

@@
 
+In this project we explore some properties of neuroevolution, the technique of applying a genetic algorithm to a neural network \cite{Garis90geneticprogramming} \cite{Sims94evolving3d}.
+
 Undoubtedly, the neural network architecture has been the star of the machine learning resurgence of the past few years.
 It's provable quality of being an universal function approximator makes it a sensible choice for various applications.
-The main algorithm used for training, backpropagation however is subject to an array of criticism from various authors \cite{some_papers_about_backprop}, and there are various attempts at improving or replacing it \cite{that_paper_where_ga_beats_backprop}.
-\todo{Add a sentence of context for \cite{Sims94evolving3d} and \cite{Garis90geneticprogramming}}
+The main algorithm used for training, backpropagation however is subject to an array of criticism from various authors \cite{zhang2007hybrid, riedmiller1993direct}, and there are various attempts at improving or replacing it \cite{montana1989training}.
 
-In this project we explore some properties of neuroevolution, the technique of applying a genetic algorithm to a neural network \cite{Garis90geneticprogramming} \cite{Sims94evolving3d}.
-A big advantage of this scheme is that for a series of different inputs a single scalar indicator of performance is sufficient in order to optimize a neural network for a certain task.
+A big advantage of the neuroevolution scheme is that for a series of different inputs a single scalar indicator of performance is sufficient in order to optimize a neural network for a certain task.
 This can for instance be given by the outcome of a game.
@@
 
-This implementation of the idea from \cite{Sims94evolving3d} uses two species that are always evaluated against the other. We evaluated
+This implementation of the idea from \cite{Sims94evolving3d} uses two species (named left and right) that are always evaluated against the other. We evaluated
 

### 2019-01-07 documentation/paper.tex
Reread and todos

@@
 Undoubtedly, the neural network architecture has been the star of the machine learning resurgence of the past few years.
-It's provable quality of being an universal function approximator makes it a sensible choice for various applications.
-The main algorithm used for training, backpropagation however is subject to an array of criticism from various authors \cite{zhang2007hybrid, riedmiller1993direct}, and there are various attempts at improving or replacing it \cite{montana1989training}.
+It's provable property of being an universal function approximator makes it a sensible choice for various applications.
+The main algorithm used for training, backpropagation however is subject to an array of criticism from authors (and even from Hinton himself), and there are various attempts at improving or replacing it \cite{zhang2007hybrid, riedmiller1993direct, montana1989training}.
 
-A big advantage of the neuroevolution scheme is that for a series of different inputs a single scalar indicator of performance is sufficient in order to optimize a neural network for a certain task.
+A big advantage of the neuroevolution scheme is that for a series of different inputs a single scalar indicator of performance is sufficient to optimize a neural network for a certain task.
 This can for instance be given by the outcome of a game.
@@
 We can see that the algorithm achieves reasonable scores against the two easiest opponents, \texttt{trainer} and \texttt{hyp014}.
-Since we were not able to get significantly better results by hyperparameter-tweaking, we theorized that the discrepancy might come from the trainers being too hard to beat by any initial network mutation, therefore providing the same zero fitness for any single useful mutation.
+Since we were not able to get significantly better results by hyperparameter-tweaking, we theorized that the discrepancy might come from the trainers being too hard to beat by any initial network mutation, providing the same zero fitness for any single useful mutation.
 Therefore we introduced a weaker, "benevolent" set of trainers that give up after scoring five points by not returning the ball anymore.
@@
 
-This implementation of the idea from \cite{Sims94evolving3d} uses two species (named left and right) that are always evaluated against the other. We evaluated
-
-\todo{Did we have two 100 sized pops here?}
+This implementation of the idea from \cite{Sims94evolving3d} uses two species (named left and right, after the side they played on in-game) that are always evaluated against the other.
+The population size was 100 for each of the species.
 
@@
 In this scenario \ref{fig:self_adapt_1_self_training_2_oldrun} can be seen battling for being the better species in the beginning, but then this seems to converge.
-A quick manual check confirms however, that they are both still actively learning.
+A quick manual look at the games confirms however, that they are both still actively learning.
 When left alone to train for 8000 generations (figure \ref{fig:self_adapt1_self_training_2}) we get the best results so far (figure \ref{fig:self_training_2}).
@@
 
-\todo{rationalize not messing around with more mutation and crossover types as we promised}
+In the proposal, we listed more possible mutation and crossover operators from \cite{montana1989training}, that we later decided not to implement.
+The main reason behind is that these extra operators were based on modifying network architecture or size.
+After the first promising test runs, we decided on a fixed size, fully-connected architecture supported by hyperparameter self-adaptation, so we would have the compute time to pursue the more interesting self-training approach instead of searching the hyperparameter space of multiple architecture operators.
+Any other connection scheme, or smaller network can be represented by setting the correct weights to zero in our bigger network.
 
@@
 \todo{train the self trained bot with reduced so we maybe get a perfect score of 12}
-\todo{we have not used too many inputs I still coudn't beat the bot which trained itself for four hours}
-\todo{acknowledgement to the compute server}
+
+We would like to thank \todo{xyz lab or something} TU-Chemnitz for letting us use their compute servers for our experiments.
 

### 2019-01-07 documentation/paper.tex
Small fixes

@@
 This can for instance be given by the outcome of a game.
-Our game of choice is Blobby Volley 2 which is an open source, simple two-dimensional volleyball game written in C++ with lua support. See Figure \ref{fig:screenshot}.
+Our game of choice is Blobby Volley 2 which is an open source, simple two-dimensional volleyball game written in C++ with lua support. Example picture in figure \ref{fig:screenshot}.
 
@@
 The game comes bundled with a number of bots with different styles, that are designed to pose a challenge to a wide range of human opponents.
-In order to define a clean ranking between them, we recorded the score of all possible lineups.
-The results can be seen in table no.
-\ref{tab:bot_vs_bot}.
+In order to define a clean ranking between them, we recorded the score of all possible lineups (figure \ref{tab:bot_vs_bot}).
 \todo{explain the table}
@@
 
-We implemented our own bot to act as an initial trainer \texttt{trainer} for the network.
-This trainer is easier to defeat.
-We made a copy of each bot where it gives up after a certain amount of touches with the ball, such that normal gameplay will be rewarded.
+We implemented our own bot marked \texttt{trainer} to act as an initial trainer for the network.
+This trainer is easier to defeat than the other built-in players.
 The simplification gives the evolutionary algorithm the opportunity to start optimizing the neural network in a valid direction, without introducing additional factors to the fitness score.

### 2019-01-07 documentation/paper.tex
More small fixes
Title
Last fixes

@@
 \date{\today}
-\title{Really Good Title}
+\title{Neuroevolution in a simple two-player game situation}
 \subtitle{Genetic Algorithms and Evolutionary Computing Project}
@@
 \label{tab:bot_vs_bot}
-\caption{Scores of bots competing against each other}
+\caption{Wins of bots competing against each other}
 \end{figure}
@@
 The input nodes got fed with the x and y positions of the player, it's relative position compared to the ball, and the velocity vector of the ball.
-The output nodes consisted of two neurons with tanh activations, one responsible for going left or right in the next timestep (if the output was below or above of threshold values $t_l = 0.49, t_r = 0.51$) and the other for jumping (above threshold $t_j = 0.7$).
+The output nodes consisted of two neurons with $\tanh$ activations, one responsible for going left or right in the next timestep (if the output was below or above of threshold values $t_l = 0.49, t_r = 0.51$) and the other for jumping (above threshold $t_j = 0.7$).
 
@@
 
-\todo{overall: not very pretty graphs <<-- Why do we decided to do self adaptation?}
+Since running the algorithm is still quite resource intensive, we wanted to focus on improving the algorithm itself instead of hyperparameter optimization.
 
-We first implemented a method similar to (1) and (4) from \cite{self_adapt}, giving each individual a mutation rate and a sigma value.
+We first implemented a self adaptation method similar to (1) and (4) from \cite{self_adapt}, giving each individual a mutation rate and a sigma value.
 These were mutated when the individual was selected for mutation.
@@
 
-\todo{too much diversity <<-- what does this mean?}
-
-In this scenario \ref{fig:self_adapt_1_self_training_2_oldrun} can be seen battling for being the better species in the beginning, but then this seems to converge.
+In this scenario (figure \ref{fig:self_adapt_1_self_training_2_oldrun}) the two species can be seen battling for higher fitness in the beginning, but then this seems to converge.
 A quick manual look at the games confirms however, that they are both still actively learning.
@@
 The only bot capable of matching this network is the strongest \texttt{reduced} bot, but it never trained against any of the built-in bots directly.
-(In fact, this network was too strong even for us humans to beat in the game.)
+(In fact, our last neural network was too strong even for us humans to beat in the game.)
 
@@
 
+A video of our two self-trained bots playing games can be found at \texttt{documentation/two_species.mp4}
+
 In the proposal, we listed more possible mutation and crossover operators from \cite{montana1989training}, that we later decided not to implement.
@@
 Any other connection scheme, or smaller network can be represented by setting the correct weights to zero in our bigger network.
+We also skipped implementing tournament-selection because of time constraints.
 
-\todo{everything we have not done}
-\todo{train the self trained bot with reduced so we maybe get a perfect score of 12}
+In the future we would like to improve on our best network by combining self train and the trainers, in order to produce an undefeated champion of the game.
+Since the project is open-source, we also plan on contributing some of our agents back to the original.
 
-We would like to thank \todo{xyz lab or something} TU-Chemnitz for letting us use their compute servers for our experiments.
+We would like to thank Faculty of Mathematics of TU-Chemnitz for letting us use their compute servers for our experiments.
 

### 2019-01-07 documentation/paper.tex

@@
 \maketitle
-\todo{think about good title}
 
@@
 The game comes bundled with a number of bots with different styles, that are designed to pose a challenge to a wide range of human opponents.
-In order to define a clean ranking between them, we recorded the score of all possible lineups (figure \ref{tab:bot_vs_bot}).
-\todo{explain the table}
+In order to define a clean ranking between them, we recorded the score of all possible lineups (figure \ref{tab:bot_vs_bot} contains the wins of bots in the row entry versus the column entry bot).
 
@@
 \label{tab:bot_vs_bot}
-\caption{Wins of bots competing against each other}
+\caption{Scores of bots competing against each other}
 \end{figure}

### 2019-01-08 documentation/paper.tex
little clean-up

@@
 Mutation is accomplished by adding Gaussian noise to a subset of all weights and biases.
-We selected an individual for mutation with a probability of $0.7$, which then had a per-neuron mutation rate of $0.5$, and a sigma of $0.01$.
+We selected an individual for mutation with a probability of $0.7$, which then had a per-neuron mutation rate of $0.5$, and a standard derivation $\sigma=0.01$.
 
@@
 
-\todo{trainer5t does not coincide on the two graphs, what happened? We get a good result against trainer5t according to the bottom graph}
-
 \section{Per-parameter self adaptation}
@@
 These bots previously worked as solid teachers.
-As can be seen on \ref{fig:change}, only the Per-individual self adaptation seems to work, there's no transfer of knowledge between the different bots for the other methods.
+As can be seen on \ref{fig:change}, only the per-individual self adaptation seems to work, there's no transfer of knowledge between the different bots for the other methods.
 
@@
 
-A video of our two self-trained bots playing games can be found at \texttt{documentation/two_species.mp4}
+A video of our two self-trained bots playing games can be found under the name \texttt{two\_species.mp4}
 
@@
 Since the project is open-source, we also plan on contributing some of our agents back to the original.
+This future work can be tracked on the GitHub repositoy in \cite{github_repo}.
 
